### 2024-11-20 paper/paper.qmd
deleted extra files, made changes in paper qmd

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-11-20 scripts/00-simulate_data.R
finished all testing scripts

@@ -1,12 +1,11 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
-  #state and party that won each division.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
-# License: MIT
-# Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
+# Purpose: Simulates a data set of the variables needed for housing affordability
+# analysis, including CPI, interest rates, housing prices, average wages, and housing supply
+# Author: Krishna Kumar
+# Date: 20 November, 2024
+# Contact: krishna.kumar@mail.utoronto.ca
+# Pre-requisites: None
+# Any other information needed? N/A
 
 
 #### Workspace setup ####
@@ -50,6 +49,9 @@ simulated_data <- data.frame(
   average_wages = average_wages,
   house_supply = house_supply
 )
+
+# Ensure the date column is of type Date
+simulated_data$date <- as.Date(simulated_data$date)
 
 # View the first few rows of the simulated data
 head(simulated_data)

### 2024-11-20 scripts/01-test_simulated_data.R

@@ -13,77 +13,87 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(testthat)
+library(dplyr)
 
-analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+## reading data
 
-# Test if the data was successfully loaded
-if (exists("analysis_data")) {
-  message("Test Passed: The dataset was successfully loaded.")
-} else {
-  stop("Test Failed: The dataset could not be loaded.")
+simulated_data <- read.csv("data/00-simulated_data/simulated_data.csv")
+
+head(simulated_data)
+
+# 1. Test: Dataset load success
+test_that("Dataset loads successfully", {
+  expect_true(exists("simulated_data") && is.data.frame(simulated_data))
+})
+
+# 2. Test: No missing values in the dataset
+test_that("No missing values in dataset", {
+  expect_true(all(!is.na(simulated_data)))
+})
+
+# 3. Test: Column names are correct
+test_that("Column names are as expected", {
+  expected_names <- c("date", "cpi", "interest_rates", "house_prices", "average_wages", "house_supply")
+  expect_equal(colnames(simulated_data), expected_names)
+})
+
+# 4. Test: Column data types are correct
+test_that("Column data types are correct", {
+  expect_true(is.Date(simulated_data$date))              # Date column
+  expect_true(is.numeric(simulated_data$cpi))            # CPI column
+  expect_true(is.numeric(simulated_data$interest_rates)) # Interest rates
+  expect_true(is.numeric(simulated_data$house_prices))   # House prices
+  expect_true(is.numeric(simulated_data$average_wages))  # Average wages
+  expect_true(is.numeric(simulated_data$house_supply))   # House supply
+})
+
+# 5. Test: Value ranges for each column
+test_that("Values fall within expected ranges", {
+  expect_true(all(simulated_data$cpi >= 200 & simulated_data$cpi <= 400))               # CPI range
+  expect_true(all(simulated_data$interest_rates >= 0 & simulated_data$interest_rates <= 10)) # Interest rates
+  expect_true(all(simulated_data$house_prices >= 200000 & simulated_data$house_prices <= 600000)) # House prices
+  expect_true(all(simulated_data$average_wages >= 20 & simulated_data$average_wages <= 50))      # Average wages
+  expect_true(all(simulated_data$house_supply >= 4 & simulated_data$house_supply <= 12))         # House supply
+})
+
+# 6. Test: Data set shape (number of rows and columns)
[…]

### 2024-11-20 scripts/02-download_data.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Downloads and saves data from FRED, needed for analysis
+# Author: Krishna Kumar
+# Date: 20 November, 2024
+# Contact: krishna.kumar@mail.utoronto.ca
+# Pre-requisites: None
+# Any other information needed? N/A
 
 
 #### Workspace setup ####

### 2024-11-20 scripts/03-clean_data.R

@@ -1,55 +1,131 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Combines all datasets and cleans them, making them ready for analysis
+# Author: Krishna Kumar
+# Date: 20 November, 2024
+# Contact: krishna.kumar@mail.utoronto.ca
+# Pre-requisites: Download all datasets according to "download_data" R script
+# Any other information needed? N/A
 
 #### Workspace setup ####
+
 library(tidyverse)
 library(readr)
+library(arrow)
 
-#### Clean data ####
-raw_data <- read_csv("data/01-raw_data/usa_00006.csv.gz")
+## Reading the datasets in this file 
 
-head(raw_data)
-tail(raw_data)
-colnames(raw_data)
+cpi_data <- read.csv("data/01-raw_data/cpi_data.csv")
+interest_rate_data <- read.csv("data/01-raw_data/interest_rate_data.csv")
+income_data <- read.csv("data/01-raw_data/income_data.csv")
+house_price_data <- read.csv("data/01-raw_data/house_price_data.csv")
+house_supply_data <- read.csv("data/01-raw_data/house_supply_data.csv")
 
-# Keep only relevant columns
-cleaned_data <- raw_data %>%
-  select(YEAR, MULTYEAR, SERIAL, STATEICP, MORTAMT1, RENTGRS, HHINCOME, VALUEH, SEX, AGE, EMPSTAT)
+## Inspecting the datasets
+
+head(cpi_data)
+tail(interest_rate_data)
+tail(income_data)
+tail(house_price_data)
+head(house_supply_data)
+
+## Converting first columns of all data sets to 'date' types
+
+cpi_data$date <- as.Date(cpi_data$date)
+interest_rate_data$date <- as.Date(interest_rate_data$date)
+income_data$date <- as.Date(income_data$date)
+house_price_data$date <- as.Date(house_price_data$date)
+house_supply_data$date <- as.Date(house_supply_data$date)
+
+## Checking if the column type conversion worked
+
[…]

### 2024-11-20 scripts/04-test_analysis_data.R

@@ -11,59 +11,69 @@
 #### Workspace setup ####
 library(tidyverse)
 library(testthat)
+library(dplyr)
+library(arrow)
 
-data <- read_csv("data/02-analysis_data/analysis_data.csv")
+analysis_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
+head(analysis_data)
 
+# Ensure date column is of type Date
+analysis_data$Date <- as.Date(analysis_data$Date)
 
-#### Test data ####
-# Test that the dataset has 151 rows - there are 151 divisions in Australia
-test_that("dataset has 151 rows", {
-  expect_equal(nrow(analysis_data), 151)
+# 1. Test: Dataset loads successfully
+test_that("Dataset loads successfully", {
+  expect_true(exists("analysis_data") && is.data.frame(analysis_data))
 })
 
-# Test that the dataset has 3 columns
-test_that("dataset has 3 columns", {
-  expect_equal(ncol(analysis_data), 3)
-})
-
-# Test that the 'division' column is character type
-test_that("'division' is character", {
-  expect_type(analysis_data$division, "character")
-})
-
-# Test that the 'party' column is character type
-test_that("'party' is character", {
-  expect_type(analysis_data$party, "character")
-})
-
-# Test that the 'state' column is character type
-test_that("'state' is character", {
-  expect_type(analysis_data$state, "character")
-})
-
-# Test that there are no missing values in the dataset
-test_that("no missing values in dataset", {
+# 2. Test: No missing values in the dataset
+test_that("No missing values in dataset", {
   expect_true(all(!is.na(analysis_data)))
 })
 
-# Test that 'division' contains unique values (no duplicates)
-test_that("'division' column contains unique values", {
-  expect_equal(length(unique(analysis_data$division)), 151)
+# 3. Test: Column names are correct
+test_that("Column names are as expected", {
+  expected_names <- c("Date", "CPI (Index)", "Interest Rate (%)", "Average Wage ($/Hour)", 
+                      "House Price (USD)", "House Supply (Month's Supply)")
+  expect_equal(colnames(analysis_data), expected_names)
 })
 
-# Test that 'state' contains only valid Australian state or territory names
-valid_states <- c("New South Wales", "Victoria", "Queensland", "South Australia", "Western Australia", 
-                  "Tasmania", "Northern Territory", "Australian Capital Territory")
[…]

### 2024-11-22 paper/paper.qmd
finished all testing scripts
almost completed data section, modified analysis data to add two new variables

@@
 author: 
-  - First author
-  - Another author
+  - Krishna Kumar
 thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
@@
 ---
-
-```{r}
-#| include: false
-#| warning: false
-#| message: false
-
-library(tidyverse)
-library(palmerpenguins)
-```
 
@@
 
-We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...
+This report uses monthly economic data derived from the Federal Reserve Economic Data (FRED) database to examine the key factors influencing housing affordability in the United States of America from March 2006 to September 2024.
+For this analysis, I have combined five data sets into one, which is then used for analysis.
+The data sets were downloaded using the `fredr` **(CITE)** package and loaded using the same library.
+The data were cleaned and analysed using the statistical programming language `R`.
+Simulating, cleaning, testing, and graphing were done with the help of the following packages: `Tidyverse`, `dplyr`, `lubridate`, `testthat`, `readr`, `arrow`, `ggplot2`, `here`, **(CITE + ADD ANY MORE YOU USED HERE) this**
 
-Overview text
+The Federal Research Economic Data provides information about the people and economy of the United States, with a goal to support economic growth, assist in making informed decisions, and enhance scientific knowledge.
+Housing affordability has emerged as a growing concern in the US, with people facing a shortage of houses and constantly increasing prices due to a variety of factors.
+These challenges, compounded by the COVID-19 pandemic, have left many families struggling to access affordable housing.
+**(CITE 1) FROM REFERENCES)** Studying this data is essential in assessing the economic factors affecting housing affordability as well as the financial wellbeing of individuals.
+The data sets were all downloaded separately using the FRED API and then combined during the data cleaning.
 
-## Measurement
+## Measurement (LEFT TO DO)
 
-Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.
+For each variable:
 
-## Outcome variables
+-   `CPI`:
 
-Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.
+-   `Interest Rate`:
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+-   `Average Wage`:
+
+-   `House Price`:
+
+-   `House Supply`:
+
+## Data Cleaning
+
+The main aspect of the data cleaning was merging all five data sets into one data set.
+Moreover, the columns were renamed to remove any informal names and two new columns were created using the existing data sets: House Prices Growth (%) and Wage Growth (%).
+This was done using the simple formula of \_\_\_\_ and to help with year over year change analysis of the house prices and wage rates.
+
+## Data Limitations and Similar Data sets
+
+While the data from FRED provides valuable insights into macroeconomic variables, it is not without limitations.
+Notably, the data on wages reflects average hourly earnings of all employees in the private sector, rather than total household income, which may overlook other forms of income, such as government transfers, investments, or multi-income households, that influence affordability.
+Additionally, the data is primarily aggregated at the national level, limiting its ability to capture trends at the regional level in housing affordability, which can be substantial across states and metropolitan areas.
+Another consideration is that some variables are seasonally adjusted while others are not.
+However, over a long period of time, like the nearly two-decades used in this analysis, these seasonal adjustments are unlikely to impact the broader trends and implications this paper aims to study.
+
+Similar data sets for all variables were readily available on the FRED database, along with other databases like the U.S.
+Census Bureau or Bureau of Labor Statistics.
+While those data sets might be able to provide regional or demographic data, the chosen variables were selected due to the monthly recording of data for a more detailed analysis.
+Despite these limitations, the FRED data sets turned out to be the best choice for an analysis fo this kind.
+
+## Variables of Interest
+
+For analysis, only the required variables' data sets were downloaded from FRED.
+This paper decided to focus on five variables to assess the problem of unaffordable housing in the US: CPI, interest rate, housing supply, median housing prices, and average wages.
+They are described below:
+
+-   `CPI`: This variable represents the Consumer Price Index for All Urban Consumers, for all items in the U.S.
+City Average, which measures the average change over time in the prices paid by urban consumers for a market basket of consumer goods and services.
+Recorded monthly, this variable is measured in an index form, for which the base period is 1982-1984 where the value of the index sat at 100.
+Change in CPI from last year's CPI is one of the most common measures used to calculate inflation.
+This variable is seasonally adjusted and looks at \~ 88% of the population.
+**(CITE 2) HERE FROM REFERENCES)**
+
+-   `Interest Rate`: This variable refers to the Federal Funds Effective Rate, which is the interest rate at which depository institutions trade federal funds with each other overnight.
+This interest rate is the central interest rate in the US, influencing other interest rates such as the prime rate (what banks charge customers with higher credit ratings) and longer term rates (mortgages, loans, savings).
+This variable is recorded monthly, represented as a percentage, and is not seasonally adjusted.
+**(CITE 3) HERE FROM REFERENCES)**
+
+-   `Average Wage`: This variable denotes the Average Hourly Earnings of All Employees in the Total Private Sector.
+It includes premium pay for overtime and late-shift work.
+This was chosen over wage rates as this measures the actual return to a worker for a set period of time.
+It excludes benefits, irregular bonuses, retroactive pay, and payroll taxes paid by employers.
+This variable is measured in US Dollars per hour, recorded monthly, and is seasonally adjusted.
+**(CITE 4) FROM REFERENCES)**
+
+-   `House Price`: This variable captures the Media Sales Price for New houses Sold int he United States.
+The media was chosen over the average prices to remove the effects of any outliers.
+This variable is recorded monthly, measured in US dollars, and is not seasonally adjusted.
+**(CITE 5) FROM REFERENCES)**
+
+-   `House Supply`: This variables measures the Monthly Supply of New Houses in the United States.
+Measured in 'month's supply,' this refers to the ratio of new houses for sale to new houses sold.
+The month's supply indicates the size of the new for-sale inventory compared to the number of new houses currently being sold.
+This variable is recorded monthly and is seasonally adjusted.
+**(CITE 6) FROM REFERENCES)**
+
+-   `House Price Growth`: This variable represents the year-over-year percentage growth in the Median Sales Price for New Houses Sold in the United States.
+It is calculated by taking the difference between the median house price at the end of the year and the beginning of the year, dividing this difference by the value at the beginning of the year, and multiplying by 100 to express the result as a percentage.
+This highlights the annual rate of change in house prices, helping to track trends in housing market inflation or deflation.
+
+-   `Wage Rate Growth`: This variable reflects the year-over-year percentage growth in the Average Hourly Earnings of All Employees in the Total Private Sector.
+The calculation follows the same approach as house price growth: the difference between wages at the end of the year and the beginning of the year is divided by the wage value at the start of the year, then multiplied by 100 to express the result as a percentage.
+This variable captures how wages change annually, offering insights into trends in earning power.
+
+### Consumer Price Index Trends Over Time
+
+```{r, INCREASE FIGWIDTH HERE}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-cpi
+#| fig-width: 10
+#| fig-height: 5
+#| fig-cap: 'Consumer Price Index (CPI) Over Time'
+
+library(ggplot2)
+library(here)
+library(arrow)
+
+analysis_data <- read_parquet(here("data", "02-analysis_data", "analysis_data.parquet"))
+
+# CPI graph
+cpi_plot <- ggplot(analysis_data, aes(x = Date, y = `CPI (Index)`)) +
+  geom_line(color = "blue", size = 1.2) + # Clear and soft blue line
+  labs(
+    x = "Year",
+    y = "CPI (Index)"
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), # Center-align and bold title
+    axis.title.x = element_text(size = 12, face = "plain", vjust = -1.5), # Spacing for x-axis title
+    axis.title.y = element_text(size = 12, face = "plain", margin = margin(r = 10)), # Add space between y-axis title and numbers
+    axis.text = element_text(size = 10, color = "gray30"), # Customize axis text color and size
+    axis.line = element_line(color = "black", size = 0.5), # Add solid black lines for x and y axes
+    axis.ticks = element_line(color = "black", size = 0.5), # Dark ticks for better visibility
+    panel.grid.major = element_line(color = "gray85", size = 0.5), # Lighten major grid lines
+    panel.grid.minor = element_blank(), # Remove minor grid lines for cleaner look
+    plot.margin = margin(10, 10, 10, 10) # Add margin around the plot
+  )
+
+# Display the plot
+cpi_plot
+
+```
+
+Figure **(CROSS REFERENCE HERE)** shows a clear pattern of a gradual increase from 2006 to 2024, with a clear spike in 2007 due to the financial crisis and a drop immediately after 2020 as a result of the COVID-19 pandemic.
+We can also see a notable acceleration in the post pandemic period CPI where the trend line becomes much steeper.
+This highlights periods of economic expansion, but also signals rising costs for consumers.
+
+Inflation, as measured by CPI, directly impacts housing affordability.
+As inflation rises, the cost of living increases, including housing prices.
+Higher inflation often prompts the Federal Reserve to raise interest rates (which we will see in later graphs), making mortgages more expensive and reducing affordability for potential home buyers.
+Furthermore, when inflation erodes purchasing power, individuals have less disposable income to allocate toward homeownership, exacerbating affordability challenges.
+Historical relationships between CPI and housing prices reveal that rising inflation can contribute to increased home prices due to higher costs for construction materials and labor.
+**(CITE 7) HERE FROM REFERENCES)**
+
+\newpage
+
+### Interest Rates Trends Over Time
 
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| warning: false
+#| message: false
 #| echo: false
+#| label: fig-interest-rate
+#| fig-width: 10
+#| fig-height: 5
+#| fig-cap: 'Federal Funds Effective Interest Rate Over Time'
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
+library(ggplot2)
+library(here)
+library(arrow)
+
+# Load the data
+analysis_data <- read_parquet(here("data", "02-analysis_data", "analysis_data.parquet"))
+
+# Create the Interest Rate graph
+interest_rate_plot <- ggplot(analysis_data, aes(x = Date, y = `Interest Rate (%)`)) +
+  geom_line(color = "red", size = 1) +
+  labs(
+    title = "Interest Rate Over Time",
+    x = "Year",
+    y = "Interest Rate (%)"
+  ) +
+  scale_y_continuous(
+    breaks = seq(0, 6, by = 1),  
+    limits = c(0, 6)             
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
+    axis.title.x = element_text(margin = margin(t = 10), face = "plain"),
+    axis.title.y = element_text(margin = margin(r = 15), face = "plain"),
+    panel.grid.major = element_line(color = "grey90"),
+    panel.grid.minor = element_blank(),
+    axis.line = element_line(color = "black")
+  )
+
+# Display the plot
+interest_rate_plot
+
 ```
 
-Talk more about it.
+Figure @fig-interest-rate shows the irregular trends of the interest rate from 2006 to 2024.
+The interest rates saw a drastic drop during the financial crisis which began in 2007.
+From 2008 to 2015, interest rates remained historically low, hovering near zero due to monetary policy responses to the Great Recession.
+A gradual increase began in 2016, reflecting economic recovery and attempts to normalize monetary policy, before another sharp decline in 2019 as the COVID-19 pandemic spurred aggressive monetary easing to support the economy.
+Post-2022, we can see a dramatic rise in interest rates, corresponding to central banks' efforts to combat inflationary pressures.
+This sharp increase represents one of the fastest monetary tightening cycles in recent history.
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+Regarding housing affordability, these trends are crucial.
+Low interest rates, such as those seen during 2008–2015 and 2020–2021, generally reduce borrowing costs, making mortgages more affordable and stimulating demand in the housing market.
+Ironically, as we will see in @fig-housing-supply, during these times when borrowing becomes cheaper, we have seen housing supply fall drastically.
+Moreover, the sharp increases in interest rates post-2022 significantly raise mortgage rates, reducing affordability for new buyers and potentially cooling housing market activity, and of course, in these times, we have seen an increase in housing supply.
+These trends underline the strong interplay between monetary policy and housing market dynamics, highlighting the importance of interest rate policies in influencing housing affordability over time
+
+### Housing Supply Over Time
 
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
 #| warning: false
 #| message: false
+#| echo: false
+#| label: fig-housing-supply
+#| fig-width: 10
+#| fig-height: 5
+#| fig-cap: 'Housing Supply Over Time'
 
-analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))
+library(ggplot2)
+library(here)
+library(arrow)
 
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
+# Define recession periods
+recessions <- data.frame(
+  start = as.Date(c("2007-12-01", "2020-03-01")),
+  end = as.Date(c("2009-06-01", "2020-04-01"))
+)
+
+# Plot with shaded recession areas
+ggplot(analysis_data, aes(x = Date, y = `House Supply (Month's Supply)`)) +
+  geom_line(color = "cornflowerblue", size = 1) +
+  geom_rect(data = recessions, inherit.aes = FALSE,
+            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
+            fill = "grey50", alpha = 0.2) +
+  labs(
+    x = "Year",
+    y = "Month's Supply"
+  ) +
   theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
+  theme(
+    axis.line.x = element_line(color = "grey80", size = 0.5),  # Add solid x-axis line
+    axis.title.x = element_text(margin = margin(t = 10)),       # Add vertical gap to x-axis label
+    axis.ticks.x = element_line(color = "grey80"),              # Add ticks to the x-axis
+    axis.ticks.y = element_line(color = "grey90"),              # Add ticks to the y-axis
+    axis.ticks.length = unit(6, "pt")                           # Adjust the length of the ticks
+  )
+
 ```
 
-Talk way more about it.
+Figure @fig-housing-supply highlights sharp fluctuations in housing supply in the past 19 years, with recessions marked by the shaded areas.
+In the aftermath of the 2007–2008 financial crisis, housing supply spiked to over 12 months, reflecting a surplus of available homes due to decreased demand and widespread foreclosures.
+However, as the economy began recovering, supply steadily declined, hitting a low point around 2013.
+This downward trend in supply coincided with robust demand, driven by low interest rates and favorable borrowing conditions.
 
-## Predictor variables
+During the COVID-19 pandemic, a combination of supply chain disruptions, labor shortages, and heightened demand for housing led to a sharp decline in supply, reaching historic lows.
+Post-2020, the supply began to rise again, likely due to reduced affordability as a result of rising interest rates (as seen in Figure @fig-interest-rate) and cooling housing market activity.
+However, it reached a high point in the first half of 2022 and then kept decreasing drastically till 2024.
 
-Add graphs, tables and text.
+\newpage
 
-Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.
+### Wage Growth vs House Prices
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-growth-rates
+#| fig-width: 10
+#| fig-height: 5
+#| fig-cap: 'Growth Rates of House Prices and Wages'
+
+library(ggplot2)
+library(here)
+library(arrow)
+library(dplyr)
+library(lubridate)
+
+# Creating a new tibble and adding a Year column and calculate yearly growth rates
+annual_growth <- analysis_data %>%
+  mutate(Year = year(Date)) %>%
+  group_by(Year) %>%
+  summarize(
+    `House Price Growth (%)` = (last(na.omit(`House Price (USD)`)) - first(na.omit(`House Price (USD)`))) / first(na.omit(`House Price (USD)`)) * 100,
+    `Wage Growth (%)` = (last(na.omit(`Average Wage ($/Hour)`)) - first(na.omit(`Average Wage ($/Hour)`))) / first(na.omit(`Average Wage ($/Hour)`)) * 100
+  )
+
+# Graphing
+
+library(ggplot2)
+
+# Plot the bar chart
+ggplot(annual_growth, aes(x = as.factor(Year))) +
+  # House Price Growth Bars
+  geom_bar(aes(y = `House Price Growth (%)`, fill = "House Prices"), 
+           stat = "identity", position = "stack", width = 0.8, alpha = 1) +
+  # Wage Growth Bars
+  geom_bar(aes(y = `Wage Growth (%)`, fill = "Wages"), 
+           stat = "identity", position = "stack", width = 0.8, alpha = 0.8) +
+  # Add a horizontal line at y = 0
+  geom_hline(yintercept = 0, color = "grey30", size = 0.5) +
+  # Customizing Fill Colors
+  scale_fill_manual(values = c("House Prices" = "#FF5733", "Wages" = "#3498DB")) +
+  # Adjust Axes and Labels
+  scale_y_continuous(
+    name = "Growth Rate (%)", 
+    breaks = seq(-10, 15, by = 5),  # Add y-axis breaks from -10% to 15% in intervals of 5%
+    limits = c(-12.5, 15)             # Set limits from -12.5% to 15%
+  ) +
+  labs(
+    x = "Year",
+    fill = "Legend"
+  ) +
+  theme_minimal() +
+  theme(
+    axis.title.x = element_text(margin = margin(t = 10)),   # Gap for x-axis title
+    axis.title.y = element_text(margin = margin(r = 15)),   # Gap for y-axis title
+    axis.line.x = element_blank(),                         # Remove default x-axis line
+    axis.line.y = element_line(color = "grey100", size = 0.5), # Keep y-axis line
+    axis.text.x = element_text(angle = 45, hjust = 1),      # Rotate x-axis labels for readability
+    legend.position = "top"
+  )
+
+```
+
+Figure @fig-growth-rates compares the annual growth rates of house prices and wages from 2006 to 2024.
+The data highlights a persistent gap between the two metrics, with house prices outpacing wage growth in 5 of the 19 years analysed.
+During the financial crisis of 2007–2008, house prices saw a sharp drop, with growth rates falling below -10%, while wage growth remained relatively stable.
+Post-2011, house prices recovered, showing strong growth from 2012 to 2016, significantly exceeding wage growth during this period.
+This divergence reflects a period of increasing housing demand, driven by low interest rates and insufficient housing supply, as seen in Figure @fig-housing-supply.
+
+Notably, wage growth has consistently lagged behind house price growth, contributing to worsening housing affordability over the years.
+For example, during the COVID-19 pandemic in 2019, while wage growth remained positive, house prices surged, reflecting the impact of supply shortages and pandemic-induced demand.
+Post-2022, a notable shift occurred with house price growthdeclining due to rising interest rates and reduced affordability, while wage growth maintained a steady pace.
+These trends emphasize the widening affordability gap and the complex interplay between wages, housing prices, and macroeconomic conditions, highlighting the challenges faced by households in balancing income growth with escalating housing costs.
 
@@
 
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
 ```
@@
 
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
 ```

### 2024-11-22 scripts/02-download_data.R

@@ -3,7 +3,9 @@
 # Author: Krishna Kumar
 # Date: 20 November, 2024
 # Contact: krishna.kumar@mail.utoronto.ca
-# Pre-requisites: None
+# Pre-requisites: Create an account on 'https://fred.stlouisfed.org/' and 
+  # generate an API key by going to 'My Account' and clicking on 'API Keys' 
+  # and then clicking on 'Request API Key.'
 # Any other information needed? N/A
 
 

### 2024-11-22 scripts/03-clean_data.R

@@ -11,6 +11,7 @@
 library(tidyverse)
 library(readr)
 library(arrow)
+library(dplyr)
 
 ## Reading the datasets in this file 
 
@@ -126,6 +127,22 @@ colnames(cleaned_data) <- c(
 colnames(cleaned_data)
 head(cleaned_data)
 
+## Creating two new columns for Wage Growth Rate and House Price Growth Rate
+
+# Calculate growth rates
+analysis_data <- analysis_data %>%
+  arrange(Date) %>%  # Ensure the data is sorted by date
+  mutate(
+    `House Price Growth (%)` = ( `House Price (USD)` / lag(`House Price (USD)`) - 1) * 100,
+    `Wage Growth (%)` = ( `Average Wage ($/Hour)` / lag(`Average Wage ($/Hour)`) - 1) * 100
+  )
+
+head(analysis_data)
+colnames(analysis_data)
+
+head(analysis_data[, c("Date", "House Price Growth (%)", "Wage Growth (%)")])
+head(analysis_data[, c("Date", "House Price (USD)", "Average Wage ($/Hour)")])
+
 ## Saving data
 
-write_parquet(cleaned_data, "data/02-analysis_data/analysis_data.parquet")
+write_parquet(analysis_data, "data/02-analysis_data/analysis_data.parquet")

### 2024-11-23 paper/paper.qmd
added to data section, halfway through model

@@
 
-# Introduction
+# Introduction {#sec-introduction}
 
@@
 
--   `CPI`: This variable represents the Consumer Price Index for All Urban Consumers, for all items in the U.S.
+-   **`CPI`:** This variable represents the Consumer Price Index for All Urban Consumers, for all items in the U.S.
 City Average, which measures the average change over time in the prices paid by urban consumers for a market basket of consumer goods and services.
@@
 
--   `Interest Rate`: This variable refers to the Federal Funds Effective Rate, which is the interest rate at which depository institutions trade federal funds with each other overnight.
+-   **`Interest Rate`:** This variable refers to the Federal Funds Effective Rate, which is the interest rate at which depository institutions trade federal funds with each other overnight.
 This interest rate is the central interest rate in the US, influencing other interest rates such as the prime rate (what banks charge customers with higher credit ratings) and longer term rates (mortgages, loans, savings).
@@
 
--   `Average Wage`: This variable denotes the Average Hourly Earnings of All Employees in the Total Private Sector.
+-   **`Average Wage`:** This variable denotes the Average Hourly Earnings of All Employees in the Total Private Sector.
 It includes premium pay for overtime and late-shift work.
@@
 
--   `House Price`: This variable captures the Media Sales Price for New houses Sold int he United States.
+-   **`House Price`:** This variable captures the Media Sales Price for New houses Sold int he United States.
 The media was chosen over the average prices to remove the effects of any outliers.
@@
 
--   `House Supply`: This variables measures the Monthly Supply of New Houses in the United States.
+-   **`House Supply`:** This variables measures the Monthly Supply of New Houses in the United States.
 Measured in 'month's supply,' this refers to the ratio of new houses for sale to new houses sold.
@@
 
--   `House Price Growth`: This variable represents the year-over-year percentage growth in the Median Sales Price for New Houses Sold in the United States.
+-   **`House Price Growth`:** This variable represents the year-over-year percentage growth in the Median Sales Price for New Houses Sold in the United States.
 It is calculated by taking the difference between the median house price at the end of the year and the beginning of the year, dividing this difference by the value at the beginning of the year, and multiplying by 100 to express the result as a percentage.
@@
 
--   `Wage Rate Growth`: This variable reflects the year-over-year percentage growth in the Average Hourly Earnings of All Employees in the Total Private Sector.
+-   **`Wage Rate Growth`:** This variable reflects the year-over-year percentage growth in the Average Hourly Earnings of All Employees in the Total Private Sector.
 The calculation follows the same approach as house price growth: the difference between wages at the end of the year and the beginning of the year is divided by the wage value at the start of the year, then multiplied by 100 to express the result as a percentage.
 This variable captures how wages change annually, offering insights into trends in earning power.
+
+-   `Affordability Index`: This variable measures the ratio of the Median Sales Price for New Houses Sold in the United States to the Average Hourly Earnings of All Employees in the Total Private Sector.
+It represents how many hours of work are required to afford a median-priced home.
+A higher value indicates reduced affordability, as workers would need to work more hours to buy a home.
 
@@
 
+### Affordability Index (CLEAN THIS UP)
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-growth-rates
+#| fig-width: 10
+#| fig-height: 5
+#| fig-cap: 'Growth Rates of House Prices and Wages'
+
+library(tidyverse)
+library(ggplot2)
+
+ggplot(analysis_data, aes(x = Date, y = `Affordability Index`)) +
+  geom_line(color = "blue") +
+  labs(x = "Year", y = "Affordability Index") +
+  theme_minimal()
+```
+
 # Model
+
+```{r}
+# Load necessary library
+library(tidyverse)
+
+# Fit a simple multiple linear regression model
+model <- lm(
+  formula = `Affordability Index` ~ `Interest Rate (%)` + `CPI (Index)`, 
+  data = analysis_data
+)
+
+# Summarize the model
+summary(model)
+
+```
 
@@
 ## Model set-up
+
+We aim to model the \*\*Affordability Index\*\*, which represents the number of hours of work required to afford a house, as a function of key economic variables such as interest rates and inflation (measured by CPI).
+
+$$y_i = y_i = \beta_0 + \beta_1 \cdot \mathrm{CPI}_i + \beta_2 \cdot \mathrm{House    Supply}_i + \epsilon_i
+$$
+
+Where:
+
+-   $y_i$ is the affordability index for observation iii, which measures the ratio of house prices to average wa ges, indicating housing affordability.
+
+-   $\beta_0$ is the intercept, representing the expected value of the affordability index when all predictors are zero.
+
+-   $\beta_1$ captures the effect of the CPI (Consumer Price Index) on the affordability index, reflecting inflation.
+
+-   $\beta_2$ captures the effect of the Interest Rate on the affordability index.
+
+-   $\epsilon_i$ represents the error term, assumed to follow a normal distribution with mean 0, accounting for unexplained variability in the affordability index.
 

### 2024-11-23 scripts/03-clean_data.R

@@ -128,6 +128,8 @@ colnames(cleaned_data)
 head(cleaned_data)
 
 ## Creating two new columns for Wage Growth Rate and House Price Growth Rate
+analysis_data <- cleaned_data
+head(analysis_data)
 
 # Calculate growth rates
 analysis_data <- analysis_data %>%
@@ -143,6 +145,16 @@ colnames(analysis_data)
 head(analysis_data[, c("Date", "House Price Growth (%)", "Wage Growth (%)")])
 head(analysis_data[, c("Date", "House Price (USD)", "Average Wage ($/Hour)")])
 
+## creating a new column for affordability index (house prices/average wage)
+
+# Add Affordability Index to the dataset
+analysis_data <- analysis_data %>%
+  mutate(`Affordability Index` = `House Price (USD)` / `Average Wage ($/Hour)`)
+
+# View the first few rows to confirm
+head(analysis_data[, c("Date", "House Price (USD)", "Average Wage ($/Hour)", "Affordability Index")])
+
+head(analysis_data)
 ## Saving data
 
 write_parquet(analysis_data, "data/02-analysis_data/analysis_data.parquet")

### 2024-11-23 paper/paper.qmd
finished model creation and modelsummary. yet to do diagnostics. also added another graph of the affordability index in the data section. also finished results section

@@
 
--   **`CPI`:** This variable represents the Consumer Price Index for All Urban Consumers, for all items in the U.S.
+-   ***CPI*****:** This variable represents the Consumer Price Index for All Urban Consumers, for all items in the U.S.
 City Average, which measures the average change over time in the prices paid by urban consumers for a market basket of consumer goods and services.
@@
 
--   **`Interest Rate`:** This variable refers to the Federal Funds Effective Rate, which is the interest rate at which depository institutions trade federal funds with each other overnight.
+-   ***Interest Rate:*** This variable refers to the Federal Funds Effective Rate, which is the interest rate at which depository institutions trade federal funds with each other overnight.
 This interest rate is the central interest rate in the US, influencing other interest rates such as the prime rate (what banks charge customers with higher credit ratings) and longer term rates (mortgages, loans, savings).
@@
 
--   **`Average Wage`:** This variable denotes the Average Hourly Earnings of All Employees in the Total Private Sector.
+-   ***Average Wage:*** This variable denotes the Average Hourly Earnings of All Employees in the Total Private Sector.
 It includes premium pay for overtime and late-shift work.
@@
 
--   **`House Price`:** This variable captures the Media Sales Price for New houses Sold int he United States.
+-   ***House Price:*** This variable captures the Media Sales Price for New houses Sold int he United States.
 The media was chosen over the average prices to remove the effects of any outliers.
@@
 
--   **`House Supply`:** This variables measures the Monthly Supply of New Houses in the United States.
+-   ***House Supply:*** This variables measures the Monthly Supply of New Houses in the United States.
 Measured in 'month's supply,' this refers to the ratio of new houses for sale to new houses sold.
@@
 
--   **`House Price Growth`:** This variable represents the year-over-year percentage growth in the Median Sales Price for New Houses Sold in the United States.
+-   ***House Price Growth:*** This variable represents the year-over-year percentage growth in the Median Sales Price for New Houses Sold in the United States.
 It is calculated by taking the difference between the median house price at the end of the year and the beginning of the year, dividing this difference by the value at the beginning of the year, and multiplying by 100 to express the result as a percentage.
@@
 
--   **`Wage Rate Growth`:** This variable reflects the year-over-year percentage growth in the Average Hourly Earnings of All Employees in the Total Private Sector.
+-   ***Wage Rate Growth:*** This variable reflects the year-over-year percentage growth in the Average Hourly Earnings of All Employees in the Total Private Sector.
 The calculation follows the same approach as house price growth: the difference between wages at the end of the year and the beginning of the year is divided by the wage value at the start of the year, then multiplied by 100 to express the result as a percentage.
@@
 
--   `Affordability Index`: This variable measures the ratio of the Median Sales Price for New Houses Sold in the United States to the Average Hourly Earnings of All Employees in the Total Private Sector.
+-   ***Affordability Index:*** This variable measures the ratio of the Median Sales Price for New Houses Sold in the United States to the Average Hourly Earnings of All Employees in the Total Private Sector.
 It represents how many hours of work are required to afford a median-priced home.
@@
 
-```{r, INCREASE FIGWIDTH HERE}
+```{r}
 #| warning: false
@@
 #| echo: false
-#| label: fig-growth-rates
+#| label: fig-affordability-index
 #| fig-width: 10
@@
 
+**TALK ABOUT ABOVE GRAPH HERE.**
+
 # Model
 
-```{r}
-# Load necessary library
-library(tidyverse)
-
-# Fit a simple multiple linear regression model
-model <- lm(
-  formula = `Affordability Index` ~ `Interest Rate (%)` + `CPI (Index)`, 
-  data = analysis_data
-)
-
-# Summarize the model
-summary(model)
-
-```
+**ADD WORDS HERE.**
 
@@
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+Here we briefly describe the multiple linear regression model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
 
@@
 
-$$y_i = y_i = \beta_0 + \beta_1 \cdot \mathrm{CPI}_i + \beta_2 \cdot \mathrm{House    Supply}_i + \epsilon_i
+$$y_i = y_i = \beta_0 + \beta_1 \cdot \mathrm{CPI}_i + \beta_2 \cdot \mathrm{House    Supply}_i + \beta_3 \cdot \mathrm{Interest Rate}_i +  \epsilon_i
 $$
@@
 
--   $y_i$ is the affordability index for observation iii, which measures the ratio of house prices to average wa ges, indicating housing affordability.
+-   $y_i$ is the affordability index for observation iii, which measures the ratio of house prices to average wages, indicating housing affordability.
 
@@
 
--   $\beta_2$ captures the effect of the Interest Rate on the affordability index.
+-   $\beta_2$ captures the effect of the Housing Supply on the affordability index.
+
+-   $\beta_3$ captures the effect of the Interest Rate on the affordability index
 
@@
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+## Model Justification
 
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
+**ADD WORDS HERE.**
 
@@
 We can use maths by including latex between dollar signs, for instance $\theta$.
+
+**ALSO ADD A DAG HERE.**
 
@@
 #| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
 #| label: tbl-modelresults
@@
 
+model1 <- readRDS(here("models/model_1_CPI.rds"))
+model2 <- readRDS(here("models/model_2_CPI_HousingSupply.rds"))
+model3 <- readRDS(here("models/model_3_Full.rds"))
+
 modelsummary::modelsummary(
   list(
-    "First model" = first_model
+    "Model 1" = model1, 
+    "Model 2" = model2, 
+    "Model 3" = model3
   ),
-  statistic = "mad",
   fmt = 2
 )
+
 ```
 
-# Discussion
+**TALK ABOUT RESULTS AND THEIR MEANING/IMPLICATIONS HERE.**
+
+# Discussion {#sec-discussion}
 
@@
 
-## Second discussion point
+## Second discussion point {#sec-second-point}
 
@@
 
-## Third discussion point
+## Third discussion point {#sec-third-point}
 

### 2024-11-23 scripts/06-model_data.R

@@ -10,10 +10,42 @@
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
+library(arrow)
+library(here)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+analysis_data <- read_parquet(here("data", "02-analysis_data", "analysis_data.parquet"))
+
+## Model 1: Only using CPI as a predictor ##
+
+model_1 <- lm(`Affordability Index` ~ `CPI (Index)`, data = analysis_data)
+
+## Save Model 1 ##
+
+saveRDS(model_1, "models/model_1_CPI.rds")
+
+
+## Model 2: Using CPI and Housing Suply as Predictors ##
+
+model_2 <- lm(`Affordability Index` ~ `CPI (Index)` + `House Supply (Month's Supply)`, data = analysis_data)
+
+## Saving Model 2 ## 
+
+saveRDS(model_2, "models/model_2_CPI_HousingSupply.rds")
+
+## Model 3: Using CPI, Housing Supply, and Interest Rate as predictors ##
+
+model_3 <- lm(`Affordability Index` ~ `CPI (Index)` + `House Supply (Month's Supply)` + `Interest Rate (%)`, data = analysis_data)
+
+## Saving Model 3 ##
+
+saveRDS(model_3, "models/model_3_Full.rds")
+
+
+
+
+
+
 
 ### Model data ####
 first_model <-

### 2024-11-24 paper/paper.qmd
finished model and model diagnostics

@@
 
-## Measurement (LEFT TO DO)
+## Measurement (PARTIALLY LEFT TO DO)
 
@@
 
--   `CPI`:
+-   ***CPI*****:** THe CPI is based on prices for food, clothing, shelter, and fuels; transportation fares; service fees (e.g., water and sewer service); and sales taxes.
+Prices are collected monthly from about 4,000 housing units and approximately 26,000 retail establishments across 87 urban areas.
+To calculate the index, price changes are averaged with weights representing their importance in the spending of the particular group.
+The index measures price changes (as a percent change) from a predetermined reference date.
 
--   `Interest Rate`:
+-   ***Interest Rate:*** The Federal Open Market Committee (FOMC) meets eight times a year to determine the federal funds target rate. These rates are then announced to the public.
 
--   `Average Wage`:
+-   ***Average Wage:*** **Have to go through website for this**
 
--   `House Price`:
+-   `House Price`: **No method of measurement given for this.**
 
--   `House Supply`:
+-   `House Supply`: **No method of measurement given for this.**
 
@@
 
-### Affordability Index (CLEAN THIS UP)
+### Affordability Index
 
@@
   geom_line(color = "blue") +
-  labs(x = "Year", y = "Affordability Index") +
-  theme_minimal()
+  labs(
+    x = "Year",
+    y = "Affordability Index"
+  ) +
+  theme_minimal() +
+  theme(
+    axis.title.x = element_text(margin = margin(t = 10)), # Adds margin above the x-axis title
+    axis.title.y = element_text(margin = margin(r = 10)), # Adds margin to the right of the y-axis title
+    axis.text.x = element_text(margin = margin(t = 5)),   # Adds margin above the x-axis numbers
+    axis.text.y = element_text(margin = margin(r = 5))    # Adds margin to the right of the y-axis numbers
+  )
+
 ```
 
-**TALK ABOUT ABOVE GRAPH HERE.**
+Figure @fig-affordability-index presents the trends in the affordability index from 2006 to 2024, illustrating fluctuations in the cost of housing relative to wages.
+The affordability index is a key metric that represents the number of work hours required to afford a house, with higher values indicating greater financial strain on individuals.
+The index shows notable volatility over time, particularly during key economic events.
+From 2008 to 2011, the index declined significantly, reflecting improved affordability during the global financial crisis, likely due to falling house prices and stabilizing wages.
+However, affordability worsened from 2012 onwards, as the index steadily increased, peaking around 2020.
+This peak coincides with rising house prices during the pandemic-driven housing boom, requiring more work hours to afford a home.
+Post-2020, the index shows a downward trend, reflecting corrections in housing prices amid rising interest rates.
 
@@
 
-**ADD WORDS HERE.**
+In this section, we develop and evaluate a series of models to understand the factors influencing housing affordability.
+The major challenge lies in selecting a model which is balanced in complexity and provides insightful results.
+We will be analyzing how CPI, housing supply, and interest rates affect the housing affordability in three separate models, adding one after the other.
+The goal of out modelling strategy is to find the effect of each of the predictor variables on the affordability index and their strengths.
 
-The goal of our modelling strategy is twofold. Firstly,...
+Because housing prices or wage rates alone can't determine affordability, we created an affordability index to better measure affordability.
+This index has a simply formula where we divide housing prices by average wages, leading to a number that represents how many hours an individual would need to work to buy a house outright.
 
-Here we briefly describe the multiple linear regression model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+We begin with a single predictor, CPI, to assess it's impact on housing affordability.
+Then, we add housing supply for model 2 and interest rates for model 3, arriving at a full model that shows us the effect of each of these variables marginally, as well as collectively.
+This step wise approach helps ensure that the model remains interpretable.
+Moreover, performance measures such as R2, Adjusted R\^2, AIC, BIC, and RMSE were used to evaluate model performance.
+
+**Here we briefly describe the multiple linear regression model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].**
 
@@
 
-**ADD WORDS HERE.**
+The linear regression model was chosen for this analysis to understand how each predictor variable affects affordability of housing.
+The affordability index is key to understanding patterns of affordability in the housing market; This is something average wages or house prices cannot determine along.
+CPI, housing supply, and interest rates all have impacts on housing affordability we expect to see a strong positive correlation between the affordability index and CPI and interest rates.
+Meaning that as these variables increase, the affordability index increases as well, indicating worsening affordability.
+However, we expect to see a negative relationship between the affordability index and housing supply.
+We also aim to measure which of these variables has a stronger impact on housing affordability.
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+An increase in interest rates increases the mortgage rates in turn and decreases housing affordability.
+An increase in CPI represents an increase in inflation, which then usually leads to an increase in housing prices.
+Lastly, a decrease in the housing supply leads to a shortage of homes and therefore, leads to increased housing prices and lower affordability.
+These factors work alone and together, to influence the financial burden on individuals.
+We decided not to choose wage growth or house price growth as predictor variables in our analysis as those would have a direct conflict with the outcome variable, and would lead to an inaccurate model.
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+We opted for a linear regression model due to its capacity to quantify the marginal, and collective, effects of each predictor (CPI, housing supply, and interest rates) on housing affordability.
+This structure is well suited to match the topic and context of our analysis.
+
+All modelling was conducted using the Base R package **(CITE HERE)**, specifically using the lm() function from the statistical package for linear regression analysis.
 
@@
 #| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
+#| tbl-cap: "Model Performance Summary showing improved fit and accuracy as Housing Supply and Interest Rates are added, with $R^2$ increasing from 0.370 in Model 1 to 0.471 in Model 3, and RMSE decreasing by over 70 points"
 #| warning: false
@@
 
-**TALK ABOUT RESULTS AND THEIR MEANING/IMPLICATIONS HERE.**
+Table @tbl-modelresults summarizes the performance metrics for our three models, with progressively added variables.
+Model 1, which only uses CPI as a predictor, achieves an $R^2$ of 0.370, indicating that CPI alone explains about 37% of the variance in the affordability index.
+Model 2 uses housing supply as an added predictor and achieves only a 0.005% increase in $R^2$ , indicating that housing supply has a very minute impact on the variation of the affordability index.
+Moreover, model 2 only sees a very small reduction in the RMSE (850.31) showing a status-quo in predictive accuracy and model fit.
+This allows us to conclude that housing supply, when combined with CPI does not produce an impact that is statistically significant.
+
+Model 3, which incorporates all three predictor variables, shows the largest improvement, with $R^2$ rising to 46.4% and RMSE dropping to 782.15, reflecting a stronger model fit.
+Moreover, model 3 also sees the lowest values of AIC and BIC that indicate that model 3 is the best model as it balances fit and complexity most effectively.
+This progression highlights the benefit of adding contextual variables like housing supply and interest rates to better capture the accuracy of the model, as well as analyzing the marginal impacts of each predictor variable.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: fig-modelresultsgraph
+#| fig-cap: "Coefficient estimates with 95% confidence intervals across three models, showing that CPI and interest rate positively affect the affordability index, reducing affordability, while an increase in housing supply improves affordability."
+#| fig-width: 10
+#| fig-height: 5
+
+library(modelsummary)
+
+modelplot(
+  list("Model 1" = model1, "Model 2" = model2, "Model 3" = model3),
+  coef_omit = "(Intercept)"
+) +
+  theme_minimal() +
+  labs(
+    x = "Estimate",
+    y = "Predictor Variables"
+  ) +
+  theme(
+    axis.title.x = element_text(margin = margin(t = 10)), 
+    axis.title.y = element_text(margin = margin(r = 10)), 
+    axis.text.x = element_text(margin = margin(t = 5)),   
+    axis.text.y = element_text(margin = margin(r = 5))    
+  )
+
+```
+
+Moreover, @fig-modelresultsgraph visually represents the coefficients of the predictors in the three models, with their corresponding confidence intervals.
+Each point represents the estimated effect size (coefficient) of a predictor, and the lines extending from each point indicate the range of uncertainty (confidence intervals).
+
+The CPI coefficients remain positive for all three models, indicating that higher inflation (measured by CPI) is associated with a higher affordability index (i.e.
+less affordable housing).
+However, we see that the coefficients for housing supply are in the negatives for both models it's used as a predictor in.
+Suggesting that an increase in housing supply improves affordability, which is consistent with our expectations.
+Lastly, the coefficient for interest rate is strongly positive, indicating that higher interest rates significantly increase the affordability index (worsening affordability), likely due to higher mortgage payments.
+The wider confidence intervals of housing supply and interest rate depict uncertainty, while the absence of a confidence interval in CPI represents more statistically precise estiamtes.
+The results we have seen from @tbl-modelresults and @fig-modelresultsgraph are completely consistent with our broad economic expectations, however, this model has room for improvement in its predictive accuracy, model fit, and certainty.
 
@@
 #| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
+#| label: fig-diagnostic-plot
+#| fig-cap: "Diagnostic Plots of Model 3 show that assumptions are largely met"
 #| fig-subcap: ["Trace plot", "Rhat plot"]
@@
 
+model1 <- readRDS(here("models/model_1_CPI.rds"))
+model2 <- readRDS(here("models/model_2_CPI_HousingSupply.rds"))
+model3 <- readRDS(here("models/model_3_Full.rds"))
+
+# Generate diagnostic plots for model1
+par(mfrow = c(2, 2)) # Set layout for 4 plots
+plot(model3)
+
 ```
+
+The diagnostic plot for Model 3 is given in @fig-diagnostic-plot and we notice the following:
+
+1.
+**Residuals vs.
+Fitted**: The residuals display a somewhat random scatter around the zero line, indicating that the linearity assumption is largely met.
+However, slight deviations suggest potential minor nonlinearity, particularly in specific ranges of fitted values.
+This indicates room for improvement but does not severely undermine the model's reliability.
+
+2.
+**Q-Q Plot**: The residuals follow a relatively straight line, indicating that the assumption of normality is generally satisfied.
+Small deviations at the tails are visible, but these are minor and unlikely to significantly impact inference or predictions.
+
+3.
+**Scale-Location Plot**: The residuals appear to be evenly spread, demonstrating that the assumption of homoscedasticity (constant variance of errors) is broadly met.
+There are slight variations in spread, but these are relatively minor and unlikely to skew results.
+
+4.
+**Residuals vs.
+Leverage**: A few points with slightly higher leverage are observed, but their Cook's distance values indicate that they are not excessively influential.
+These points warrant attention but do not dominate the overall model.
 

### 2024-11-24 paper/paper.qmd
removed posterior checks since i made a linear regression model

@@
 
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-```
-
 ## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
 
@@
 Fitted**: The residuals display a somewhat random scatter around the zero line, indicating that the linearity assumption is largely met.
-However, slight deviations suggest potential minor nonlinearity, particularly in specific ranges of fitted values.
+However, slight deviations suggest potential minor non linearity, particularly in specific ranges of fitted values.
 This indicates room for improvement but does not severely undermine the model's reliability.

### 2024-11-24 paper/paper.qmd
finished data section

@@
 
--   ***CPI*****:** THe CPI is based on prices for food, clothing, shelter, and fuels; transportation fares; service fees (e.g., water and sewer service); and sales taxes.
+-   ***CPI*****:** The CPI is based on prices for food, clothing, shelter, and fuels; transportation fares; service fees (e.g., water and sewer service); and sales taxes.
 Prices are collected monthly from about 4,000 housing units and approximately 26,000 retail establishments across 87 urban areas.
@@
 Moreover, the columns were renamed to remove any informal names and two new columns were created using the existing data sets: House Prices Growth (%) and Wage Growth (%).
-This was done using the simple formula of \_\_\_\_ and to help with year over year change analysis of the house prices and wage rates.
+This was done using the simple formula of (House Price 2 - House Price 1)/House Price 1, calculating the percentage change.
+The same was done for average wages.
+This allowed us to analyse the growth rates year over year and get a sense of how fast each variable was growing through time.
+Additionally, another column was created titled 'Affordability Index.' This column was created by dividing the values of House Prices by the Average Wage in the same rows.
+This gave us an outcome of the number of hours an individual would have to work to buy a house outright, serving as a good indicator of housing affordability, especially when looking at the data through a long-term lens.
 
@@
 
-**Here we briefly describe the multiple linear regression model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].**
+Here we briefly describe the multiple linear regression model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
 
@@
 
-We aim to model the \*\*Affordability Index\*\*, which represents the number of hours of work required to afford a house, as a function of key economic variables such as interest rates and inflation (measured by CPI).
+We aim to model the Affordability Index, which represents the number of hours of work required to purchast a house outright, as a function of key economic variables such as interest rates and inflation (measured by CPI).
 
@@
 ```{r}
+#| warning: false
+#| message: false
 #| echo: false
@@
 
-## First discussion point {#sec-first-point}
+## Overview: What Is Done In This Paper?  {#sec-overview}
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+This paper investigates housing affordability in the United States over the period of 2006–2024 by developing a Affordability Index.
+The index measures the number of hours of work required to purchase a median-priced home and serves as a metric for gauging the accessibility of home ownership across time, especially when analysed through a long-term lens.
+Using multiple linear regression, this paper examines the influence of key economic variables—Consumer Price Index (CPI), housing supply, and interest rates—on housing affordability.
+These factors were chosen because they capture inflationary pressures, supply constraints, and monetary policy, which are all fundamental drivers of housing market dynamics.
 
-## Second discussion point {#sec-second-point}
+To conduct the analysis, five data sets from FRED were downloaded, compiled, and processed to align temporal and unit measurements.
+The model specifications were carefully designed to progressively evaluate the contribution of each predictor, starting with CPI as a standalone variable, followed by the inclusion of housing supply, and finally incorporating interest rates.
+This approach allowed for a clearer understanding of how these variables individually and collectively influence affordability.
+The results offer critical insights into how shifts in macroeconomic factors affect home-buying accessibility and shed light on the broader housing market trends over the last two decades.
 
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+## Key Findings {#sec-key-findings}
 
-## Third discussion point {#sec-third-point}
+### Relationship Between CPI (Inflation) and the Affordability Index  {#sec-CPI}
 
-## Weaknesses and next steps
+The positive and statistically significant relationship between CPI and the affordability index confirms that higher inflation leads to reduced affordability.
+Inflation, as measured by the CPI, directly correlates with rising housing costs because construction materials, labor, and overall living expenses increase.
+As CPI rises, potential homebuyers face higher costs not only for housing but also for other goods and services, reducing their disposable income for mortgage payments.
+This finding highlights the economic strain placed on individuals during periods of inflation and highlights the need for policies that mitigate inflationary pressures in housing markets.
 
-Weaknesses and next steps should also be included.
+### Relationship Between Housing Supply and the Affordability Index  {#sec-housing-supply}
+
+The negative coefficient associated with housing supply indicates that an increase in available housing improves affordability.
+When housing supply grows, competition among buyers decreases, stabilizing or reducing home prices.
+This relationship aligns with economic theory, which states that greater supply relative to demand leads to price reductions.
+However, the relatively modest effect size, from our model, suggests that while increasing supply is a necessary condition for improving affordability, it may not be sufficient on its own without addressing broader structural issues, such as zoning restrictions and construction costs.
+However, this could also be due to the limitations of the model.
+
+### Relationship Between Interest Rates and the Affordability Index
+
+Interest rates exert the most substantial impact on affordability among the predictors.
+The strong positive coefficient indicates that rising interest rates significantly worsen affordability by increasing mortgage costs, as interest rates affect other rates like the mortgage rates as well.
+For instance, a higher interest rate raises monthly payments, reducing the number of households that qualify for home loans.
+Moreover, looking at this from the perspective of the potential hoome buyer, it limits their choices and they are often forced to decide against buying a home and renting.
+This finding emphasizes the critical role of monetary policy in shaping housing market accessibility.
+Policymakers must carefully consider the downstream effects of interest rate hikes, especially in a high-inflation environment where affordability is already strained.
+While interest rates are often adjusted to combat the problem of growing inflation in an economy, its subsequent effects that affect the quality of life of citizens need to be considered as well.
+
+## Limitations {#sec-limitations}
+
+### Data Aggregation {#sec-data-aggregation}
+
+The data sets were aggregated at the national level, which limits the analysis' ability to capture regional variations in housing affordability.
+Housing markets in urban centers like San Francisco or New York operate under vastly different conditions than rural markets, and these distinctions are obscured in a national-level analysis.
+Studying the differences in housing affordability in rural vs urban areas as well as east vs west coast would be a very interesting, relevant, and topical concept to understand the true nature of housing affordability in the US.
+Future studies should aim to disaggregate data to better account for these geographic differences and further the analysis.
+
+### Model Fit and Scope {#sec-model-fit-and-scope}
+
+The adjusted $R^2$ values for the models range between 37% and 46%, indicating that the predictors explain less than half of the variation in the Affordability Index.
+This suggests that other unobserved factors—such as demographic shifts, wage inequality, government policies and support, regional income distributions, and renting and the supply of apartment complexes—play significant roles in determining housing affordability.
+For instance, regional wage disparities or the availability of housing subsidies could significantly influence affordability but were not included in this analysis.
+
+### Exclusion of Other Influential Variables {#sec-exclusion-of-variables}
+
+As the adjusted $R^2$ values for the model were relatively low, we can decipher that there are several other factors that account for the remaining variance in the affordability index.
+Key factors such as household savings rates, employment stability, government intervention programs (e.g., tax credits or housing subsidies), rental properties, apartment complexes, condos, planning restrictions, and regional variations were not included due to data limitations.
+These factors likely play a critical role in shaping housing affordability but remain unexplored in this analysis.
+This, in no way, is a negative point of this paper, rather, it serves as a suggestion for extension of this analysis in further studies where we could understand the problem of housing affordability even more accurately.
+
+## Future Steps
+
+### Incorporating Additional Predictors
+
+Expanding the model to include variables such as wage growth, population density, regional housing prices and income, and construction costs could enhance its explanatory power.
+For instance, wage growth relative to inflation could provide deeper insights into purchasing power, while construction costs could shed light on the supply-side challenges of the housing market.
+Conducting state- or city-level analyses would allow for a more granular understanding of housing affordability trends.
+This approach could highlight regional disparities and inform localized policy interventions.
+For example, comparing affordability in high-cost metropolitan areas to that in rural regions could reveal targeted strategies for improving housing access and government intervention.
+
+### **Exploring Non-Linear Relationships**
+
+Future studies could employ non-linear modeling techniques, such as machine learning or interaction models like a Bayesian model, to capture more complex dynamics between variables.
+For example, the interaction between interest rates and inflation may have compounding effects on affordability that are not captured by a linear framework.
+
+### **Evaluating Policy Interventions**
+
+Research focusing on the effectiveness of government programs, such as housing subsidies, rent control, or first-time homebuyer incentives, could provide actionable recommendations for policymakers.
+Evaluating these interventions in the context of affordability trends would help determine which policies have the most significant impact.
+
+In conclusion, this paper contributes to the broader understanding of housing affordability by analyzing the interplay of inflation, housing supply, and interest rates.
+The findings emphasize the critical role of macroeconomic variables in shaping housing accessibility and highlight areas where policymakers can intervene to mitigate affordability challenges.
+While the analysis has its limitations, it lays the groundwork for future research to build more comprehensive models and address the pressing issue of housing affordability in the United States.
 
@@
 #| fig-cap: "Diagnostic Plots of Model 3 show that assumptions are largely met"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
+#| fig-width: 10
+#| fig-height: 5
 

### 2024-11-25 paper/paper.qmd
finished intro, abstract, title, subtitle

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Analyzing Housing Affordability in the United States: The Role of Inflation, Interest Rates, and Supply"
+subtitle: "Rising Inflation and Interest Rates Worsen Affordability, While Increasing Housing Supply Offers Relief"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper investigates housing affordability in the United States from 2006 to 2024 by constructing an Affordability Index that measures the number of work hours required to purchase a median-priced home outright.
+Using a multiple linear regression model, we analyze the relationships between housing affordability and three key macroeconomic variables: Consumer Price Index (CPI), housing supply, and interest rates.
+Our findings reveal that rising inflation and interest rates significantly reduce affordability, while increased housing supply mitigates this effect.
+These insights highlight the complex interplay of economic forces shaping the housing market and emphasize the importance of stabilizing inflation, increasing supply, and maintaining balanced monetary policy to improve housing accessibility for U.S.
+households."
 format: pdf
@@
 
-Overview paragraph
+Housing affordability is a growing issue in the United States, with significant implications for economic stability and social well-being of individuals.
+Over the past two decades, escalating home prices, inflation, and shifting interest rates have exacerbated housing affordability issues, leaving many households unable to achieve home ownership.
+As policymakers seek solutions to these challenges, understanding the factors that drive housing affordability has become increasingly critical.
+This paper addresses this issue by focusing on the interplay of key macroeconomic variables—Consumer Price Index (CPI), housing supply, and interest rates—and their collective influence on housing affordability over time.
 
-Estimand paragraph
+To capture the complexity of this issue, we develop an Affordability Index, a measure representing the number of work hours required to purchase a median-priced home outright.
+This index provides a clear, intuitive way to assess affordability trends int he long run and evaluate the impact of inflation, housing supply, and monetary policy on home ownership accessibility.
+Using multiple linear regression, we analyze how these factors interact and contribute to changes in affordability from 2006 to 2024.
+By systematically building and comparing three models, we examine the power of each predictor (marginally and collectively) and evaluate their significance in shaping affordability dynamics.
 
-Results paragraph
+The findings of this paper reveal that inflation and interest rates significantly reduce housing affordability (i.e.
+increase the affordability index), while an increase in housing supply improves it.
+Rising CPI drives up the cost of living and housing prices, decreasing purchasing power.
+Similarly, higher interest rates inflate mortgage costs, further straining affordability.
+Conversely, increased housing supply alleviates upward pressure on prices, demonstrating its critical role in improving affordability.
+These results offer valuable insights for policymakers, highlighting the need for measures that stabilize inflation, increase housing stock, and balance monetary policy to enhance housing accessibility.
 
-Why it matters paragraph
+The remainder of this paper is organized as follows.
+Section @sec-data presents the data sources, variable descriptions, and exploratory analysis.
+Section @sec-model outlines the model details, setup, and justification.
+Section @sec-results highlights the key findings from the models and visualizations.
+Finally, Section @sec-discussion delves into the broader implications of these findings, addresses limitations, and suggests directions for future research.
+Lastly Section **(FIGURE OUT HOW TO REFERENCE APPENDIX HERE)** presents the model diagnostics and a survey, sampling, and observation method that would be ideal for data of the kind we used.
 
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
+## Estimand
+
+The estimand in this paper is the Affordability Index, representing the hours of work required to purchase a median-priced home outright.
+This measure allows us to quantify and analyze the impact of macroeconomic variables on affordability in a clear and meaningful way, especially when the analysis focuses on the long run.
+By focusing on this estimand, we contribute to a deeper understanding of housing market dynamics and provide actionable insights for addressing affordability challenges.
 
@@
 
-# Model
+# Model {#sec-model}
 
@@
 
-# Results
+# Results {#sec-results}
 

### 2024-11-25 paper/paper.qmd
finished survey, sampling, and observation appendix

@@
 Finally, Section @sec-discussion delves into the broader implications of these findings, addresses limitations, and suggests directions for future research.
-Lastly Section **(FIGURE OUT HOW TO REFERENCE APPENDIX HERE)** presents the model diagnostics and a survey, sampling, and observation method that would be ideal for data of the kind we used.
+Moreover, Sections @sec-model-details and @sec-ideal-meth of the appendix present the model diagnostics and a survey, sampling, and observation methodology that would be ideal for nature of data that we have used.
 
@@
 
+## Idealized Methodology {#sec-ideal-meth}
+
+### Limitations with FRED Methodology
+
+The Federal Reserce Economic Data's (FRED) data sets, while comprehensive in providing macroeconomic indicators, have several limitations when assessing housing affordability.
+First, it lacks granularity at the household level, failing to capture individual variations in financial burdens or regional disparities.
+Second, FRED focuses on objective metrics like house prices, wages, and economic variables, but omits subjective indicators such as perceived affordability or financial stress felt by individuals.
+Third, the data is often aggregated, potentially obscuring critical subpopulations like low-income households or first-time home buyers.
+These gaps emphasize the need for complementary data collection methods that can provide a more holistic understanding of housing affordability.
+
+### Overview
+
+To address the gaps identified in the FRED data sets and collection methods, we propose a comprehensive survey methodology aimed at capturing a clearer view of housing affordability in the United States.
+This approach integrates best practices from existing housing surveys, such as the Canada Mortgage and Housing Corporation's (CMHC) Social and Affordable Housing Survey **\[CITE 9) FROM REFERENCES\]**, and emphasizes both objective metrics and subjective experiences.
+The methodology focuses on diverse sampling techniques, inclusive recruitment strategies, and solid data validation processes to ensure the collection of reliable and representative data.
+
+### Sampling Method
+
+To ensure that the survey captures a wide range of experiences and perspectives regarding housing affordability, we propose a **stratified multi-stage sampling design**.
+This design breaks down the population into smaller, more manageable groups and ensures that each group is properly represented in the final sample.
+Here’s how it works:
+
+1.
+**Stratification**: The first step divides the population into groups, or "*strata*," based on key characteristics such as geographic region, income level, type of housing (renter or homeowner), and urban or rural residence.
+For example, one stratum might consist of low-income renters in urban areas, while another might consist of middle-income homeowners in rural areas.
+By dividing the population in this way, we ensure that all important subgroups are included in the survey.
+
+2.
+**Primary Sampling Units (PSUs)**: Within each stratum, large areas such as counties or cities are chosen as primary sampling units.
+These are selected using a method called "probability proportional to size" (PPS), which means areas with larger populations are more likely to be included.
+For example, If Los Angeles has a population of 4 million and a small rural area has a population of 40,000, the probability of selecting Los Angeles as a PSU is 100 times greater than that of the rural county.
+This is done in order to keep the probabilities of selection equal to the population sizes to give each area a proportinoally equal opportunity and representation.
+
+3.
+**Secondary Sampling Units (SSUs)**: After the PSUs are chosen, smaller areas within them—such as neighborhoods—are selected.
+Again, PPS sampling is used to ensure that larger neighborhoods have a better chance of being included, due to the higher sampling frame.
+For example, within Los Angeles, densely populated neighborhoods like Downtown LA might have a higher probability of being chosen than sparsely populated suburban neighborhoods.
+However, suburban neighborhoods still have a chance to be included to ensure a balanced representation.
+
+4.  **Addressing Rural/Small Town Representation**:\
+    While PPS is efficient, it risks underrepresenting rural areas as they often have lower populations and hence, a lower chance of being selected.
+These areas often face distinct housing affordability challenges.
+To mitigate this:
+
+    -   **Separate Strata for Rural Areas:** A separate strata will be constructed for rural areas and areas with smaller populations, from which PSUs and SSUs can be selected.
+We did mention this in the stratification section, but we must emphasize that we aim for equal representation of all areas and demographic types.
+
+    -   **Oversampling Rural Areas**: Rural PSUs can be oversampled within their strata, ensuring sufficient representation despite their smaller population sizes.
+
+    <!-- -->
+
+    -   **Weighted Analysis**: Post-survey, weights can be applied to adjust for the lower selection probability of rural areas.
+For example, responses from a small rural neighborhood might be weighted more heavily than those from a densely populated city.
+
+    -   **Quota Sampling**: Quotas can be set to guarantee that a minimum percentage of the sample comes from rural PSUs, ensuring balance between urban and rural perspectives.
+
+5.
+**Household Selection**: Finally, individual households are randomly selected from the chosen neighborhoods.
+This ensures that the survey captures a variety of household experiences within each subgroup.
+
+This process creates a sample that represents the diversity of the U.S.
+population while also allowing for deeper insights into specific groups.
+For example, by oversampling low-income households or areas with high housing costs, the survey can provide detailed data about the populations most affected by housing affordability challenges.
+This method is inspired by similar approaches used in large-scale housing surveys, such as the CMHC's Social and Affordable Housing Survey, which ensures a representative sample across diverse regions and demographics **\[CITE 9) FROM REFERENCES\].**
+
+### Recruitment Strategy
+
+A well-designed recruitment strategy is crucial for ensuring a diverse and representative sample for the survey.
+To achieve this, a **mixed-mode approach** will be used, employing multiple methods to contact participants and collect responses.
+This strategy not only maximizes response rates but also ensures that no group is excluded due to accessibility or technological barriers.
+To further maximize response rates and participation, we will hope to include a small compensation (small cash amount, gift cards, etc) for completing the survey, given that we can gather sufficient funding.
+Here’s how the process works:
+
+-   **Mail Invitations**:\
+    The first step in reaching participants is through mailed invitations.
+These physical letters will provide an introduction to the survey, explaining its purpose, importance, and instructions for participation.
+Including a personal touch like mail ensures that even households without internet access or those less tech-savvy are aware of the survey.
+For instance, older adults who may not regularly use the internet are more likely to respond to mailed materials.
+
+-   **Online Surveys**:\
+    Online surveys will serve as the primary method for data collection because they are cost-effective, quick, and accessible to most people.
+Participants can simply follow a link from the mailed invitation or QR code to access the survey on their computers or smartphones.
+Online surveys also make it easier to process and analyze data efficiently, reducing both time and human error.
+This method is particularly effective for tech-savvy demographics, such as younger adults.
+The platform for these surveys will be Google Forms.
+
+-   **Telephone Interviews**:\
+    To accommodate households that lack internet access or prefer not to participate online, telephone interviews will be conducted.
+Trained interviewers will call participants and guide them through the survey, ensuring that they can share their responses comfortably.
+This approach is particularly helpful for older adults, rural households, and low-income groups who may face barriers to online participation.
+It ensures that no one is left out due to lack of internet or unfamiliarity with digital tools.
+All necessary steps will be taken to identify the household and individuals correctly to ensure full accuracy in data collection.
+
+-   **In-Person Interviews**:\
+    In areas with low response rates or where other methods prove ineffective, in-person interviews will be conducted.
+Trained interviewers will visit selected households to collect responses face-to-face.
+This method is particularly useful in remote areas or among populations that might distrust online or phone methods.
+For example, face-to-face interactions often lead to higher response rates among communities where personal rapport is valued.
+
+This mixed-mode approach is inspired by the Canadian Mortgage and Housing Corporation’s (CMHC) survey methodology, which emphasizes using diverse data collection techniques to enhance data quality and fully represent all types of individuals and households.
+It ensures that every demographic—urban or rural, young or old, digitally connected or not—is represented in the survey results.
+
+### Data Validation
+
+Ensuring the accuracy and reliability of collected data is essential. The following validation measures will be implemented:
+
+-   **Consistency Checks**: Survey responses will be examined for logical consistency, such as verifying that reported housing costs align with stated income levels.
+
+-   **Duplicate Detection**: Measures will be in place to identify and remove duplicate responses, using unique identifiers like address or phone number.
+
+-   **Non-Response Analysis**: Patterns of non-response will be analyzed to assess potential biases, and weighting adjustments will be applied to account for underrepresented groups.
+
+-   **Data Imputation**: For missing or incomplete responses, statistical imputation methods will be used to estimate values, maintaining the integrity of the data set.
+
+These validation techniques are consistent with standard practices in housing surveys, ensuring the collection of high-quality data suitable for robust analysis.
+
+### Conclusion
+
+By integrating comprehensive sampling strategies, inclusive recruitment methods, and rigorous data validation processes, this idealized survey methodology aims to provide a nuanced understanding of housing affordability in the United States.
+Drawing on established practices from surveys like those conducted by the CMHC, this approach addresses the limitations of existing datasets and offers valuable insights for policymakers and researchers seeking to address housing affordability challenges.
+While most measures of housing affordability focus on the quantitative factors, we aim to create a balance and bring in the essential qualitative factors needed to really understand how an individual or a household feels about their financial situation and housing affordability.
+For example, we plan to include questions about whether respondents feel comfortable meeting their mortgage or rent payments, and if they feel secure in their ability to handle unexpected financial shocks, such as medical bills or job loss.
+
 \newpage

### 2024-11-25 paper/paper.qmd
finished creating survey

@@
 
-Research focusing on the effectiveness of government programs, such as housing subsidies, rent control, or first-time homebuyer incentives, could provide actionable recommendations for policymakers.
+Research focusing on the effectiveness of government programs, such as housing subsidies, rent control, or first-time home buyer incentives, could provide actionable recommendations for policymakers.
 Evaluating these interventions in the context of affordability trends would help determine which policies have the most significant impact.
@@
 
-# Model details {#sec-model-details}
-
-## Diagnostics
+## Model Diagnostics {#sec-model-details}
 
@@
 
+The survey can be found at this link: <https://forms.gle/xLyZsGiEEDcwPnH66>
+
+## Copy of the Survey
+
 \newpage

### 2024-11-26 paper/paper.qmd
created DAG, finished measurements and adding survey copy to appendix

@@
 For this analysis, I have combined five data sets into one, which is then used for analysis.
-The data sets were downloaded using the `fredr` **(CITE)** package and loaded using the same library.
-The data were cleaned and analysed using the statistical programming language `R`.
-Simulating, cleaning, testing, and graphing were done with the help of the following packages: `Tidyverse`, `dplyr`, `lubridate`, `testthat`, `readr`, `arrow`, `ggplot2`, `here`, **(CITE + ADD ANY MORE YOU USED HERE) this**
+The data sets were downloaded using the `fredr` [@FRED] package and loaded using the same library.
+The data were cleaned and analysed using the statistical programming language `R` [@citeR].
+Simulating, cleaning, testing, and graphing were done with the help of the following packages: `Tidyverse` [@tidyverse], `dplyr` [@dplyr], `lubridate` [@lubridate], `testthat` [@testthat], `readr` [@readr], `arrow` [@arrow], `ggplot2` [@ggplot2], `here` [@here], `DiagrammeR` [@diagram], \``DiagrammeRsvg` [@DiagrammeRsvg], `rsvg` [@rsvg], and `modelsummary` [@modelsummary].
 
-The Federal Research Economic Data provides information about the people and economy of the United States, with a goal to support economic growth, assist in making informed decisions, and enhance scientific knowledge.
+The Federal Research Economic Data provides informati on about the people and economy of the United States, with a goal to support economic growth, assist in making informed decisions, and enhance scientific knowledge.
 Housing affordability has emerged as a growing concern in the US, with people facing a shortage of houses and constantly increasing prices due to a variety of factors.
-These challenges, compounded by the COVID-19 pandemic, have left many families struggling to access affordable housing.
-**(CITE 1) FROM REFERENCES)** Studying this data is essential in assessing the economic factors affecting housing affordability as well as the financial wellbeing of individuals.
+These challenges, compounded by the COVID-19 pandemic, have left many families struggling to access affordable housing [@1].
+Studying this data is essential in assessing the economic factors affecting housing affordability as well as the financial well-being of individuals.
 The data sets were all downloaded separately using the FRED API and then combined during the data cleaning.
 
-## Measurement (PARTIALLY LEFT TO DO)
-
-For each variable:
+## Measurement
 
@@
 
--   ***Average Wage:*** **Have to go through website for this**
+-   ***Average Wage:*** This is measured through the Current Employment Statistics (CES) program, which surveys approximately 119,000 businesses and government agencies monthly.
+Employers report total gross payroll and hours worked for all private-sector employees during the pay period including the 12th of the month.
+Average Hourly Earnings are calculated by dividing total payroll by total hours worked, excluding irregular bonuses or benefits, and aggregated using the NAICS classification system to provide an accurate measure of wage trends across industries.
 
--   `House Price`: **No method of measurement given for this.**
+-   ***House Price***: This is measured through data collected by the U.S.
+Census Bureau and the Department of Housing and Urban Development (HUD).
+This data is based on surveys of new single-family houses sold, with the price representing the middle value when all houses sold during a given period are ordered by price.
+The calculation excludes multifamily units and focuses on single-family homes, ensuring the measure reflects trends in the mid-range of housing markets without being skewed by extreme high or low prices.
 
--   `House Supply`: **No method of measurement given for this.**
+-   ***House Supply***: This is measured as the ratio of the number of new houses available for sale to the number of new houses sold, expressed in months.
+Data is collected through the New Residential Sales Survey, conducted jointly by the U.S.
+Census Bureau and HUD, which tracks both housing inventory and sales rates.
+This measure reflects how long the current inventory of new houses would last at the current sales pace, assuming no new inventory is added.
+The measure is calculated by dividing the total inventory of new homes available for sale by the current monthly sales rate.
+For example, if there are 400,000 houses for sale and 40,000 houses are sold in a month, the monthly supply is \$400,000/40,000 = 10 months.
+A higher value suggests excess inventory (a buyer's market), while a lower value indicates limited inventory (a seller's market).
 
@@
 Change in CPI from last year's CPI is one of the most common measures used to calculate inflation.
-This variable is seasonally adjusted and looks at \~ 88% of the population.
-**(CITE 2) HERE FROM REFERENCES)**
+This variable is seasonally adjusted and looks at \~ 88% of the population [@2].
 
@@
 This interest rate is the central interest rate in the US, influencing other interest rates such as the prime rate (what banks charge customers with higher credit ratings) and longer term rates (mortgages, loans, savings).
-This variable is recorded monthly, represented as a percentage, and is not seasonally adjusted.
-**(CITE 3) HERE FROM REFERENCES)**
+This variable is recorded monthly, represented as a percentage, and is not seasonally adjusted [@3].
 
@@
 It excludes benefits, irregular bonuses, retroactive pay, and payroll taxes paid by employers.
-This variable is measured in US Dollars per hour, recorded monthly, and is seasonally adjusted.
-**(CITE 4) FROM REFERENCES)**
+This variable is measured in US Dollars per hour, recorded monthly, and is seasonally adjusted [@4].
 
@@
 The media was chosen over the average prices to remove the effects of any outliers.
-This variable is recorded monthly, measured in US dollars, and is not seasonally adjusted.
-**(CITE 5) FROM REFERENCES)**
+This variable is recorded monthly, measured in US dollars, and is not seasonally adjusted [@5].
 
@@
 The month's supply indicates the size of the new for-sale inventory compared to the number of new houses currently being sold.
-This variable is recorded monthly and is seasonally adjusted.
-**(CITE 6) FROM REFERENCES)**
+This variable is recorded monthly and is seasonally adjusted [@6].
 
@@
 
-Figure **(CROSS REFERENCE HERE)** shows a clear pattern of a gradual increase from 2006 to 2024, with a clear spike in 2007 due to the financial crisis and a drop immediately after 2020 as a result of the COVID-19 pandemic.
+Figure @fig-cpi shows a clear pattern of a gradual increase from 2006 to 2024, with a clear spike in 2007 due to the financial crisis and a drop immediately after 2020 as a result of the COVID-19 pandemic.
 We can also see a notable acceleration in the post pandemic period CPI where the trend line becomes much steeper.
@@
 Higher inflation often prompts the Federal Reserve to raise interest rates (which we will see in later graphs), making mortgages more expensive and reducing affordability for potential home buyers.
-Furthermore, when inflation erodes purchasing power, individuals have less disposable income to allocate toward homeownership, exacerbating affordability challenges.
-Historical relationships between CPI and housing prices reveal that rising inflation can contribute to increased home prices due to higher costs for construction materials and labor.
-**(CITE 7) HERE FROM REFERENCES)**
+Furthermore, when inflation erodes purchasing power, individuals have less disposable income to allocate toward home ownership, exacerbating affordability challenges.
+Historical relationships between CPI and housing prices reveal that rising inflation can contribute to increased home prices due to higher costs for construction materials and labor [@7].
 
@@
 
-Here we briefly describe the multiple linear regression model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+Here we briefly describe the multiple linear regression model used to investigate... Background details and diagnostics are included in @sec-model-details of the Appendix.
 
@@
 
-We aim to model the Affordability Index, which represents the number of hours of work required to purchast a house outright, as a function of key economic variables such as interest rates and inflation (measured by CPI).
+We aim to model the Affordability Index, which represents the number of hours of work required to purchase a house outright, as a function of key economic variables such as interest rates and inflation (measured by CPI).
 
@@
 
-All modelling was conducted using the Base R package **(CITE HERE)**, specifically using the lm() function from the statistical package for linear regression analysis.
+All modelling was conducted using the Base R package, specifically using the lm() function from the statistical package for linear regression analysis.
 
-**ALSO ADD A DAG HERE.**
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-DAG
+#| fig-width: 10
+#| fig-height: 5
+#| fig-cap: 'Factors influencing the Affordability Index shown as direct contributors'
+
+library(here)
+
+# Load and display the saved PNG
+knitr::include_graphics(here("paper", "dag.png"))
+
+```
 
@@
 #| warning: false
+
+library(modelsummary)
 
@@
 Model 2 uses housing supply as an added predictor and achieves only a 0.005% increase in $R^2$ , indicating that housing supply has a very minute impact on the variation of the affordability index.
-Moreover, model 2 only sees a very small reduction in the RMSE (850.31) showing a status-quo in predictive accuracy and model fit.
+Moreover, model 2 only sees a very small reduction in the RMSE (850.31) showing a status quo in predictive accuracy and model fit.
 This allows us to conclude that housing supply, when combined with CPI does not produce an impact that is statistically significant.
@@
 Inflation, as measured by the CPI, directly correlates with rising housing costs because construction materials, labor, and overall living expenses increase.
-As CPI rises, potential homebuyers face higher costs not only for housing but also for other goods and services, reducing their disposable income for mortgage payments.
+As CPI rises, potential home buyers face higher costs not only for housing but also for other goods and services, reducing their disposable income for mortgage payments.
 This finding highlights the economic strain placed on individuals during periods of inflation and highlights the need for policies that mitigate inflationary pressures in housing markets.
@@
 
-The Federal Reserce Economic Data's (FRED) data sets, while comprehensive in providing macroeconomic indicators, have several limitations when assessing housing affordability.
+The Federal Reserve Economic Data's (FRED) data sets, while comprehensive in providing macroeconomic indicators, have several limitations when assessing housing affordability.
 First, it lacks granularity at the household level, failing to capture individual variations in financial burdens or regional disparities.
 Second, FRED focuses on objective metrics like house prices, wages, and economic variables, but omits subjective indicators such as perceived affordability or financial stress felt by individuals.
-Third, the data is often aggregated, potentially obscuring critical subpopulations like low-income households or first-time home buyers.
+Third, the data is often aggregated, potentially obscuring critical sub populations like low-income households or first-time home buyers.
 These gaps emphasize the need for complementary data collection methods that can provide a more holistic understanding of housing affordability.
@@
 To address the gaps identified in the FRED data sets and collection methods, we propose a comprehensive survey methodology aimed at capturing a clearer view of housing affordability in the United States.
-This approach integrates best practices from existing housing surveys, such as the Canada Mortgage and Housing Corporation's (CMHC) Social and Affordable Housing Survey **\[CITE 9) FROM REFERENCES\]**, and emphasizes both objective metrics and subjective experiences.
+This approach integrates best practices from existing housing surveys, such as the Canada Mortgage and Housing Corporation's (CMHC) Social and Affordable Housing Survey [@8], and emphasizes both objective metrics and subjective experiences.
 The methodology focuses on diverse sampling techniques, inclusive recruitment strategies, and solid data validation processes to ensure the collection of reliable and representative data.
@@
 4.  **Addressing Rural/Small Town Representation**:\
-    While PPS is efficient, it risks underrepresenting rural areas as they often have lower populations and hence, a lower chance of being selected.
+    While PPS is efficient, it risks under representing rural areas as they often have lower populations and hence, a lower chance of being selected.
 These areas often face distinct housing affordability challenges.
@@
 
-    -   **Oversampling Rural Areas**: Rural PSUs can be oversampled within their strata, ensuring sufficient representation despite their smaller population sizes.
-
-    <!-- -->
+    -   **Oversampling Rural Areas**: Rural PSUs can be over sampled within their strata, ensuring sufficient representation despite their smaller population sizes.
 
@@
 For example, by oversampling low-income households or areas with high housing costs, the survey can provide detailed data about the populations most affected by housing affordability challenges.
-This method is inspired by similar approaches used in large-scale housing surveys, such as the CMHC's Social and Affordable Housing Survey, which ensures a representative sample across diverse regions and demographics **\[CITE 9) FROM REFERENCES\].**
+This method is inspired by similar approaches used in large-scale housing surveys, such as the CMHC's Social and Affordable Housing Survey, which ensures a representative sample across diverse regions and demographics [@8].
 
@@
 
+**Housing Affordability Survey 2024**
+
+**Introduction:** Thank you for participating in the Housing Affordability Survey.
+This survey aims to understand how individuals and households perceive and experience housing affordability in the United States.
+Your responses will help in assessing the financial and social factors contributing to housing affordability challenges.
+
+Participation in this survey is entirely voluntary, and you can choose to withdraw at any time without any consequences.
+The information collected will be used solely for research purposes.
+Your responses will remain confidential and will only be reported in aggregate form.
+
+**Contact Information:** For questions regarding this survey or the methodology used, please reach out to:
+
+-   **Krishna Kumar**, University of Toronto. Email: [krishna.kumar\@mail.utoronto.ca](mailto:krishna.kumar@mail.utoronto.ca)
+
+### Section 1: Housing Situation
+
+1.   **What is your current housing situation?**
+
+    -   Own a home with a mortgage
+
+    -   Own a home without a mortgage
+
+    -   Rent a home/apartment
+
+    -   Live with family/friends without paying rent
+
+    -   Other (please specify): \_\_\_\_\_\_\_
+
+2.   **How much of your monthly income goes toward housing costs (mortgage/rent, utilities, etc.)?**
+
+    -   Less than 20%
+
+    -   20%-30%
+
+    -   30%-50%
+
+    -   More than 50%
+
+3.   **Do you consider your housing costs to be affordable?**
+
+    -   Yes
+
+    -   No
+
+    -   Other: \_\_\_\_\_\_\_\_
+
+4.  **How would you describe the stability of your current housing situation?**
+
+    -   Very stable
+
+    -   Somewhat stable
+
+    -   Somewhat unstable
+
+    -   Very unstable
+
+### Section 2: Financial and Social Factors
+
+5.   **Have you experienced difficulty paying for housing in the past year?**
+
+    -   Yes
+
+    -   No
+
+    -   Other: \_\_\_\_\_\_\_\_
+
+6.   **If you answered “Yes” to the previous question, what was the primary reason?**
+
+    -   Job loss or reduced income
+
+    -   Increased housing costs
+
+    -   Unexpected expenses
+
+    -   Other: \_\_\_\_\_\_\_\_
+
+7.   **Please respond according to the following statement: My current income adequately covers my housing expenses.**
+
+    -   Strongly agree
+
+    -   Somewhat agree
+
+    -   Neutral
+
+    -   Somewhat disagree
+
+    -   Strongly disagree
+
+8.   **How often do you save money after covering all your monthly expenses, including housing?**
+
+    -   Every month
+
+    -   Most months
+
+    -   Rarely
+
+    -   Never
+
+9.   **If you own a home, have you refinanced or attempted to refinance your mortgage in the past year?**
+
+    -   Yes
+
+    -   No
+
+    -   Not applicable
+
+### Section 3: Perceptions of Housing Affordability
+
+10.  **On a scale of 1 to 5, how would you rate the overall affordability of housing in your area?**
+
+    -   1 (Not affordable at all)
+
+    -   2
+
+    -   3
+
+    -   4
+
+    -   5 (Very affordable)
+
+11.  **Do you feel your local government is addressing housing affordability issues effectively?**
+
+    -   Yes
+
+    -   No
+
+    -   Unsure
+
+12.  **Would you say that you can comfortably afford your housing?**
+
+    -   Yes
+
+    -   No
+
+    -   Other: \_\_\_\_\_\_\_
+
+13.  **What are the biggest challenges to housing affordability in your opinion? (Select up to 3)**
+
+    -   High property prices
+
+    -   Rising rents
+
+    -   Insufficient housing supply
+
+    -   Low wages
+
+    -   Lack of government support
+
+    -   Other: \_\_\_\_\_\_\_
+
+14. **How much does housing quality (size, condition, location) factor into your perception of affordability?**
+
+    -   Not at all
+
+    -   A little
+
+    -   Somewhat
+
+    -   A great deal
+
+### Section 4: Demographics
+
+15. **What is your age range? (years)**
+
+    -   18-24
+
+    -   25-34
+
+    -   35-44
+
+    -   45-54
+
+    -   55-64
+
+    -   65 or older
+
+    -   Prefer not to say
+
+16. **What gender do you identify as?**
+
+    -   Male
+
+    -   Female
+
+    -   Non-binary
+
+    -   Prefer not to say
+
+17.  **Which of the following best describes your race or ethnicity?**
+
+    -   White or Caucasian
+
+    -   Black or African America
+
+    -   Hispanic or Latino of any race
+
+    -   Asian
+
+    -   Other or multiple races
+
+18. **What is the highest level of education you have completed?**
+
+    -   Less than high school
+
+    -   High school diploma or GED
+
+    -   Some college, no degree
+
+    -   Associate degree
+
+    -   Bachelor’s degree
+
+    -   Graduate or professional degree
+
+    -   Prefer not to say
+
+19. **What is your current employment status?**
+
+    -   Employed full-time
+
+    -   Employed part-time
+
+    -   Self-employed
+
+    -   Unemployed
+
+    -   Student
+
+    -   Retired
+
+    -   Prefer not to say
+
+**End Message:** Thank you for completing this survey! Your responses are invaluable in understanding housing affordability challenges across the United States.
+
 \newpage

### 2024-11-26 scripts/05-create-DAG.R

@@ -1,0 +1,39 @@
+#### Preamble ####
+# Purpose: Creates a directed acrylic graph showcasing the model set up
+# Author: Krishna Kumar
+# Date: 20 November, 2024
+# Contact: krishna.kumar@mail.utoronto.ca
+# Pre-requisites: None
+# Any other information needed? N/A
+
+#### Workspace setup ####
+#install.packages("DiagrammeRsvg")
+#install.packages("rsvg")
+
+library(DiagrammeR)
+library(DiagrammeRsvg)
+library(rsvg)
+
+# Define the causal model
+causal_model <- "digraph {
+  # Define nodes
+  node [shape = plaintext]
+    AffordabilityIndex [label = 'Affordability Index']
+    CPI [label = 'CPI']
+    InterestRate [label = 'Interest Rate']
+    HousingSupply [label = 'Housing Supply']
+
+  # Define directed edges (causal relationships)
+  edge []
+    CPI -> AffordabilityIndex
+    InterestRate -> AffordabilityIndex
+    HousingSupply -> AffordabilityIndex
+}"
+
+# Render and export the graph
+graph <- grViz(causal_model)
+
+# Save the graph as a PNG file
+export_svg(graph) |>
+  charToRaw() |>
+  rsvg::rsvg_png("paper/dag.png", width = 1000)

### 2024-11-26 paper/paper.qmd
updated preambles and added final touches to the qmd file

@@
 
+## Data Cleaning
+
+The main aspect of the data cleaning was merging all five data sets into one data set.
+Moreover, the columns were renamed to remove any informal names and two new columns were created using the existing data sets: House Prices Growth (%) and Wage Growth (%).
+This was done using the simple formula of (House Price 2 - House Price 1)/House Price 1, calculating the percentage change.
+The same was done for average wages.
+This allowed us to analyse the growth rates year over year and get a sense of how fast each variable was growing through time.
+Additionally, another column was created titled 'Affordability Index.' This column was created by dividing the values of House Prices by the Average Wage in the same rows.
+This gave us an outcome of the number of hours an individual would have to work to buy a house outright, serving as a good indicator of housing affordability, especially when looking at the data through a long-term lens.
+
+## Data Limitations and Similar Data sets
+
+While the data from FRED provides valuable insights into macroeconomic variables, it is not without limitations.
+Notably, the data on wages reflects average hourly earnings of all employees in the private sector, rather than total household income, which may overlook other forms of income, such as government transfers, investments, or multi-income households, that influence affordability.
+Additionally, the data is primarily aggregated at the national level, limiting its ability to capture trends at the regional level in housing affordability, which can be substantial across states and metropolitan areas.
+Another consideration is that some variables are seasonally adjusted while others are not.
+However, over a long period of time, like the nearly two-decades used in this analysis, these seasonal adjustments are unlikely to impact the broader trends and implications this paper aims to study.
+
+Similar data sets for all variables were readily available on the FRED database, along with other databases like the U.S.
+Census Bureau or Bureau of Labor Statistics.
+While those data sets might be able to provide regional or demographic data, the chosen variables were selected due to the monthly recording of data for a more detailed analysis.
+Despite these limitations, the FRED data sets turned out to be the best choice for an analysis fo this kind.
+
+## Variables of Interest
+
+For analysis, only the required variables' data sets were downloaded from FRED.
+This paper decided to focus on five variables to assess the problem of unaffordable housing in the US: CPI, interest rate, housing supply, median housing prices, and average wages.
+They are described below:
+
+-   ***CPI*****:** This variable represents the Consumer Price Index for All Urban Consumers, for all items in the U.S.
+City Average, which measures the average change over time in the prices paid by urban consumers for a market basket of consumer goods and services.
+Recorded monthly, this variable is measured in an index form, for which the base period is 1982-1984 where the value of the index sat at 100.
+Change in CPI from last year's CPI is one of the most common measures used to calculate inflation.
+This variable is seasonally adjusted and looks at \~ 88% of the population [@2].
+
+-   ***Interest Rate:*** This variable refers to the Federal Funds Effective Rate, which is the interest rate at which depository institutions trade federal funds with each other overnight.
+This interest rate is the central interest rate in the US, influencing other interest rates such as the prime rate (what banks charge customers with higher credit ratings) and longer term rates (mortgages, loans, savings).
+This variable is recorded monthly, represented as a percentage, and is not seasonally adjusted [@3].
+
+-   ***Average Wage:*** This variable denotes the Average Hourly Earnings of All Employees in the Total Private Sector.
+It includes premium pay for overtime and late-shift work.
+This was chosen over wage rates as this measures the actual return to a worker for a set period of time.
+It excludes benefits, irregular bonuses, retroactive pay, and payroll taxes paid by employers.
+This variable is measured in US Dollars per hour, recorded monthly, and is seasonally adjusted [@4].
+
+-   ***House Price:*** This variable captures the Media Sales Price for New houses Sold int he United States.
+The media was chosen over the average prices to remove the effects of any outliers.
+This variable is recorded monthly, measured in US dollars, and is not seasonally adjusted [@5].
+
+-   ***House Supply:*** This variables measures the Monthly Supply of New Houses in the United States.
+Measured in 'month's supply,' this refers to the ratio of new houses for sale to new houses sold.
+The month's supply indicates the size of the new for-sale inventory compared to the number of new houses currently being sold.
+This variable is recorded monthly and is seasonally adjusted [@6].
+
+-   ***House Price Growth:*** This variable represents the year-over-year percentage growth in the Median Sales Price for New Houses Sold in the United States.
+It is calculated by taking the difference between the median house price at the end of the year and the beginning of the year, dividing this difference by the value at the beginning of the year, and multiplying by 100 to express the result as a percentage.
+This highlights the annual rate of change in house prices, helping to track trends in housing market inflation or deflation.
+
+-   ***Wage Rate Growth:*** This variable reflects the year-over-year percentage growth in the Average Hourly Earnings of All Employees in the Total Private Sector.
+The calculation follows the same approach as house price growth: the difference between wages at the end of the year and the beginning of the year is divided by the wage value at the start of the year, then multiplied by 100 to express the result as a percentage.
+This variable captures how wages change annually, offering insights into trends in earning power.
+
+-   ***Affordability Index:*** This variable measures the ratio of the Median Sales Price for New Houses Sold in the United States to the Average Hourly Earnings of All Employees in the Total Private Sector.
+It represents how many hours of work are required to afford a median-priced home.
+A higher value indicates reduced affordability, as workers would need to work more hours to buy a home.
+
 ## Measurement
@@
 
-## Data Cleaning
-
-The main aspect of the data cleaning was merging all five data sets into one data set.
-Moreover, the columns were renamed to remove any informal names and two new columns were created using the existing data sets: House Prices Growth (%) and Wage Growth (%).
-This was done using the simple formula of (House Price 2 - House Price 1)/House Price 1, calculating the percentage change.
-The same was done for average wages.
-This allowed us to analyse the growth rates year over year and get a sense of how fast each variable was growing through time.
-Additionally, another column was created titled 'Affordability Index.' This column was created by dividing the values of House Prices by the Average Wage in the same rows.
-This gave us an outcome of the number of hours an individual would have to work to buy a house outright, serving as a good indicator of housing affordability, especially when looking at the data through a long-term lens.
-
-## Data Limitations and Similar Data sets
-
-While the data from FRED provides valuable insights into macroeconomic variables, it is not without limitations.
-Notably, the data on wages reflects average hourly earnings of all employees in the private sector, rather than total household income, which may overlook other forms of income, such as government transfers, investments, or multi-income households, that influence affordability.
-Additionally, the data is primarily aggregated at the national level, limiting its ability to capture trends at the regional level in housing affordability, which can be substantial across states and metropolitan areas.
-Another consideration is that some variables are seasonally adjusted while others are not.
-However, over a long period of time, like the nearly two-decades used in this analysis, these seasonal adjustments are unlikely to impact the broader trends and implications this paper aims to study.
-
-Similar data sets for all variables were readily available on the FRED database, along with other databases like the U.S.
-Census Bureau or Bureau of Labor Statistics.
-While those data sets might be able to provide regional or demographic data, the chosen variables were selected due to the monthly recording of data for a more detailed analysis.
-Despite these limitations, the FRED data sets turned out to be the best choice for an analysis fo this kind.
-
-## Variables of Interest
-
-For analysis, only the required variables' data sets were downloaded from FRED.
-This paper decided to focus on five variables to assess the problem of unaffordable housing in the US: CPI, interest rate, housing supply, median housing prices, and average wages.
-They are described below:
-
--   ***CPI*****:** This variable represents the Consumer Price Index for All Urban Consumers, for all items in the U.S.
-City Average, which measures the average change over time in the prices paid by urban consumers for a market basket of consumer goods and services.
-Recorded monthly, this variable is measured in an index form, for which the base period is 1982-1984 where the value of the index sat at 100.
-Change in CPI from last year's CPI is one of the most common measures used to calculate inflation.
-This variable is seasonally adjusted and looks at \~ 88% of the population [@2].
-
--   ***Interest Rate:*** This variable refers to the Federal Funds Effective Rate, which is the interest rate at which depository institutions trade federal funds with each other overnight.
-This interest rate is the central interest rate in the US, influencing other interest rates such as the prime rate (what banks charge customers with higher credit ratings) and longer term rates (mortgages, loans, savings).
-This variable is recorded monthly, represented as a percentage, and is not seasonally adjusted [@3].
-
--   ***Average Wage:*** This variable denotes the Average Hourly Earnings of All Employees in the Total Private Sector.
-It includes premium pay for overtime and late-shift work.
-This was chosen over wage rates as this measures the actual return to a worker for a set period of time.
-It excludes benefits, irregular bonuses, retroactive pay, and payroll taxes paid by employers.
-This variable is measured in US Dollars per hour, recorded monthly, and is seasonally adjusted [@4].
-
--   ***House Price:*** This variable captures the Media Sales Price for New houses Sold int he United States.
-The media was chosen over the average prices to remove the effects of any outliers.
-This variable is recorded monthly, measured in US dollars, and is not seasonally adjusted [@5].
-
--   ***House Supply:*** This variables measures the Monthly Supply of New Houses in the United States.
-Measured in 'month's supply,' this refers to the ratio of new houses for sale to new houses sold.
-The month's supply indicates the size of the new for-sale inventory compared to the number of new houses currently being sold.
-This variable is recorded monthly and is seasonally adjusted [@6].
-
--   ***House Price Growth:*** This variable represents the year-over-year percentage growth in the Median Sales Price for New Houses Sold in the United States.
-It is calculated by taking the difference between the median house price at the end of the year and the beginning of the year, dividing this difference by the value at the beginning of the year, and multiplying by 100 to express the result as a percentage.
-This highlights the annual rate of change in house prices, helping to track trends in housing market inflation or deflation.
-
--   ***Wage Rate Growth:*** This variable reflects the year-over-year percentage growth in the Average Hourly Earnings of All Employees in the Total Private Sector.
-The calculation follows the same approach as house price growth: the difference between wages at the end of the year and the beginning of the year is divided by the wage value at the start of the year, then multiplied by 100 to express the result as a percentage.
-This variable captures how wages change annually, offering insights into trends in earning power.
-
--   ***Affordability Index:*** This variable measures the ratio of the Median Sales Price for New Houses Sold in the United States to the Average Hourly Earnings of All Employees in the Total Private Sector.
-It represents how many hours of work are required to afford a median-priced home.
-A higher value indicates reduced affordability, as workers would need to work more hours to buy a home.
+\newpage
 
@@
 #| fig-height: 5
-#| fig-cap: 'Consumer Price Index (CPI) Over Time'
+#| fig-cap: 'Trend in Consumer Price Index (CPI) from 2008 to 2024, highlighting a steady rise in inflation with accelerated growth post-2020.'
 
@@
 #| fig-height: 5
-#| fig-cap: 'Federal Funds Effective Interest Rate Over Time'
+#| fig-cap: 'Historical trend of interest rates from 2008 to 2024, showcasing significant declines post-2008 financial crisis and sharp increases in recent years to combat inflation.'
 
@@
 #| fig-height: 5
-#| fig-cap: 'Housing Supply Over Time'
+#| fig-cap: 'Monthly housing supply trends from 2008 to 2024, highlighting peaks during the 2008 financial crisis and the COVID-19 pandemic, followed by declines as markets adjusted.'
 
@@
 #| fig-height: 5
-#| fig-cap: 'Growth Rates of House Prices and Wages'
+#| fig-cap: 'Comparative growth rates of house prices and wages from 2006 to 2024, showing periods where house price increases significantly outpaced wage growth, particularly during housing market recoveries and economic expansions.'
 
@@
 #| fig-height: 5
-#| fig-cap: 'Growth Rates of House Prices and Wages'
+#| fig-cap: 'Trends in the Housing Affordability Index from 2006 to 2024, highlighting periods of increased affordability during economic downturns and decreased affordability during housing market peaks.'
 
@@
 
-# Additional data details
-
-## Model Diagnostics {#sec-model-details}
+# Model Diagnostics {#sec-model-details}
 
@@
 #| label: fig-diagnostic-plot
-#| fig-cap: "Diagnostic Plots of Model 3 show that assumptions are largely met"
+#| fig-cap: "Diagnostic Plots of Model 3 show that assumptions are largely met, with slight deviations"
 #| fig-width: 10
@@
 
-## Idealized Methodology {#sec-ideal-meth}
-
-### Limitations with FRED Methodology
+# Idealized Methodology {#sec-ideal-meth}
+
+## Limitations with FRED Methodology
 
@@
 
-### Overview
+## Overview
 
@@
 
-### Sampling Method
+## Sampling Method
 
@@
 
-### Recruitment Strategy
+## Recruitment Strategy
 
@@
 
-### Data Validation
+## Data Validation
 
@@
 
-### Conclusion
+## Conclusion
 
@@
 
-### Section 1: Housing Situation
+Section 1: Housing Situation
 
@@
 
-### Section 2: Financial and Social Factors
+Section 2: Financial and Social Factors
 
@@
 
-### Section 3: Perceptions of Housing Affordability
+Section 3: Perceptions of Housing Affordability
 
@@
 
-### Section 4: Demographics
+Section 4: Demographics
 

### 2024-11-26 scripts/01-test_simulated_data.R

@@ -8,7 +8,7 @@
 # Pre-requisites: 
   # - The `tidyverse` package must be installed and loaded
   # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
+# Any other information needed? Make sure you are in the `housing_affordability` rproj
 
 
 #### Workspace setup ####

### 2024-11-26 scripts/03-clean_data.R

@@ -3,7 +3,7 @@
 # Author: Krishna Kumar
 # Date: 20 November, 2024
 # Contact: krishna.kumar@mail.utoronto.ca
-# Pre-requisites: Download all datasets according to "download_data" R script
+# Pre-requisites: Download all data sets according to "download_data" R script
 # Any other information needed? N/A
 
 #### Workspace setup ####

### 2024-11-26 scripts/04-test_analysis_data.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 26 September 2024 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Tests the analysis data we will be using for our analysis
+# Author: Krishna Kumar
+# Date: 20 November, 2024
+# Contact: krishna.kumar@mail.utoronto.ca
+# Pre-requisites: none
+# Any other information needed? N/A
 
 
 #### Workspace setup ####

### 2024-11-26 scripts/05-create-DAG.R

@@ -3,10 +3,11 @@
 # Author: Krishna Kumar
 # Date: 20 November, 2024
 # Contact: krishna.kumar@mail.utoronto.ca
-# Pre-requisites: None
+# Pre-requisites: un-comment and install necessary packages as outlined below, if not already installed
 # Any other information needed? N/A
 
 #### Workspace setup ####
+
 #install.packages("DiagrammeRsvg")
 #install.packages("rsvg")
 

### 2024-11-26 scripts/06-model_data.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Runs the linear regression models for our data and saves them as .rd files
+# Author: Krishna Kumar
+# Date: 20 November, 2024
+# Contact: krishna.kumar@mail.utoronto.ca
+# Pre-requisites: none
+# Any other information needed? N/A
 
 
 #### Workspace setup ####

### 2024-11-26 paper/paper.qmd
updated readme, deleted unnecessary files, finished sketches
added thanks message

@@
   - Krishna Kumar
-thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
+thanks: "Code and data are available at: [https://github.com/krishnak30/US_housing_affordability](https://github.com/krishnak30/US_housing_affordability)."
 date: today
@@
 
+```         
 -   Own a home with a mortgage
@@
 -   Other (please specify): \_\_\_\_\_\_\_
+```
 
@@
 
+```         
 -   Less than 20%
@@
 -   More than 50%
+```
 
@@
 
+```         
 -   Yes
@@
 -   Other: \_\_\_\_\_\_\_\_
+```
 
@@
 
+```         
 -   Job loss or reduced income
@@
 
+```         
 -   Strongly agree
@@
 -   Strongly disagree
+```
 
@@
 
+```         
 -   Every month
@@
 -   Never
+```
 
@@
 -   Not applicable
+```
 
@@
 
+```         
 -   1 (Not affordable at all)
@@
 -   5 (Very affordable)
+```
 
@@
 -   Unsure
+```
 
@@
 -   Other: \_\_\_\_\_\_\_
+```
 
@@
 
+```         
 -   High property prices
@@
 
+```         
 -   White or Caucasian
@@
 -   Other or multiple races
+```
 
